---
title: "03_Leaflitter_cierra_eda"
author: "Cierra Lizer"
date: "2025-06-05"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```
**Note: I noticed I've been graphing some by week and some by day of year, I will need to choose for continuity sake**


Set up file paths etc. --> this should also indicate where you can find these files!
```{r}
path.google <- "~/Google Drive/My Drive"
path.litter <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data")
path.figs <- file.path(path.litter, "figures") # where we shoudl save some figures
path.save <- file.path(path.litter, "LeafLitterData_Clean_forArchiving") # Where we shoudl save the data
```

Using a formatting theme consistent with what Meghan has done
```{r}
theme_base <-   theme(panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      panel.border = element_rect(fill="white", colour = "black", linewidth=0.7),
                      axis.title.x = element_text(margin = margin(t = 10, b=5), size=14),
                      axis.title.y = element_text(margin = margin(l = 5, r=5), size=14),
                      axis.text.x= element_text(margin = margin(t = 10), size=12),
                      axis.text.y=element_text(margin = margin(r = 10), size=12),
                      axis.ticks.length=unit(-0.3, "cm"),
                      # axis.ticks.margin=unit(0.5, "cm"),
                      axis.ticks = element_line(colour = "black", linewidth = 0.4))


# Setting a consistent color scheme across all graphs
plotOrder <- c("B-127", "U-134", "N-115", "HH-115")
ewPlotColors <- c("#1B9E77","#D95F02", "#7570B3", "#E7298A")
names(ewPlotColors) = plotOrder
ewPlotColors
```

**FIX THIS TO GO THROUGH FILE PATHS**
```{r}
dir(path.save)
all_cleaned_files <- list.files(path.save, pattern = "\\.csv$", full.names = TRUE)
if (length(all_cleaned_files) == 0) { stop(...) }
datLeafLitter <- map_df(all_cleaned_files, read_csv)
datLeafLitter$plot <- factor(datLeafLitter$plot, levels = plotOrder)
```
```{r}
datLeafLitter$year <- lubridate::year(datLeafLitter$date_collection)
datLeafLitter$yday <- lubridate::yday(datLeafLitter$date_collection)
datLeafLitter$week <- lubridate::week(datLeafLitter$date_collection)
```

Making a filter to have just the leaf tissue type
```{r}
leaf_data <- datLeafLitter %>%
  filter(tissue == "leaf")
```


____________________________________________________________________________________________________________


Histogram to show distribution of Leaf Mass
```{r}
ggplot(leaf_data, aes(x = mass_g)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7) +
  labs(
    x = "Leaf Mass (g)",
    y = "Frequency"
  ) +
   scale_fill_manual(values=ewPlotColors) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

Boxplot to show year to year variation in leaf mass
```{r}
ggplot(leaf_data, aes(x=factor(year), y=mass_g, fill = factor(year))) +
  geom_boxplot() +
  labs(
    x="Year",
    y="Leaf Mass (g)",
    fill="Year"
  )+
  scale_fill_manual(values=ewPlotColors) +
  theme_bw()
```

Boxplot of leaf mass by year but changed to have a square root scale to better see interquartile range
```{r}
ggplot(leaf_data, aes(x = factor(year), y = mass_g, fill = factor(year))) +
  geom_boxplot() +
scale_y_sqrt() +
  labs(
    x = "Year",
    y = "Leaf Mass (g^1/2)",
    fill = "Year"
  ) +
  scale_fill_manual(values=ewPlotColors) +
  theme_bw()
```
Boxplot to show week to week variation in leaf mass (across all years)
```{r}
ggplot(leaf_data, aes(x = factor(week), y = mass_g, fill = factor(week))) +
  geom_boxplot() +
  labs(
    x = "Week",
    y = "Leaf Mass (g)",
    fill = "Week" # Label for the fill legend
  ) +
  scale_fill_manual(values=ewPlotColors) +
  theme_bw()
```
Boxplot of leaf mass by week w/ sqrt scale
```{r}
ggplot(leaf_data, aes(x = factor(week), y = mass_g, fill = factor(week))) +
  geom_boxplot() +
  scale_y_sqrt() +
  labs(
    x = "Week",
    y = "Leaf Mass (g^1/2)",
    fill = "Week" # Label for the fill legend
  ) +
  scale_fill_manual(values=ewPlotColors) +
  theme_bw()
```

Tissue Type by Week
```{r}
tissue_mass_by_week <- datLeafLitter %>%
  group_by(week, tissue) %>%
  summarise(total_mass_g = sum(mass_g, na.rm = TRUE)) %>%
  ungroup() # Always ungroup after summarizing

ggplot(tissue_mass_by_week, aes(x = factor(week), y = total_mass_g, fill = tissue)) +
  geom_col() +
  labs(
    x = "Week",
    y = "Total Mass (g)",
    fill = "Tissue Type"
  )
```

Leaf mass by genus by week
```{r}
leaf_mass_by_genus_week <-  %>%
  filter(tissue == "leaf") %>%
  group_by(year, week, genus) %>%
  summarise(total_leaf_mass_g = sum(mass_g, na.rm = TRUE)) %>%
  ungroup()

ggplot(leaf_mass_by_genus_week, aes(x = factor(week), y = total_leaf_mass_g, fill = genus)) +
  geom_col() + 
  facet_wrap(~ year, scales = "free_x")
  labs(
    title = "Total Leaf Mass by Genus and Week",
    x = "Week of Year",
    y = "Total Leaf Mass (g)",
    fill = "Genus"
  )
```
Calculate 50% Cumulative Leaf Fall Day for each year
```{r}
leaf_fall_timing50 <- leaf_data %>%
  group_by(year, yday) %>%
  summarise(daily_leaf_mass = sum(mass_g, na.rm = TRUE), .groups = 'drop_last') %>%
  mutate(cumulative_leaf_mass = cumsum(daily_leaf_mass),
         annual_total_leaf_mass = max(cumulative_leaf_mass, na.rm = TRUE)) %>%
  ungroup() %>%
  
# Filter to find the first day where cumulative mass crosses 50% threshold
  filter(cumulative_leaf_mass >= 0.5 * annual_total_leaf_mass) %>%
  group_by(year) %>%
  slice_min(order_by = yday, n = 1) %>% 
  select(year, leaf_fall_doy_50percent = yday) %>%
  ungroup()


print(leaf_fall_timing50)
```

Can do the same by plot
```{r}
leaf_fall_timing_by_plot <- leaf_data %>%
  group_by(year, plot, yday) %>%
  summarise(daily_leaf_mass = sum(mass_g, na.rm = TRUE), .groups = 'drop_last') %>%
  mutate(cumulative_leaf_mass = cumsum(daily_leaf_mass),
         annual_plot_total_leaf_mass = max(cumulative_leaf_mass, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(cumulative_leaf_mass >= 0.5 * annual_plot_total_leaf_mass) %>%
  group_by(year, plot) %>%
  slice_min(order_by = yday, n = 1) %>% 
  select(year, plot, leaf_fall_plot_50percent = yday) %>% 
  ungroup()

print(leaf_fall_timing_by_plot)
```
**Note: we can play around with what percentage we use to find the best fit**

```{r}
ggplot(data=leaf_data)+
  facet_grid(year~plot)+
  geom_point(aes(x=yday, y=mass_g, color=factor(plot)))+
  stat_summary(geom="line", aes(x=yday, y=mass_g), fun="mean")+
  geom_vline(data=leaf_fall_timing_by_plot, aes(xintercept = leaf_fall_plot_50percent), linetype="dashed")+ labs(x="day of year", y="mass (g)")+
   scale_color_manual(values=ewPlotColors) +
  theme_bw()  
```



___________________________________________________________________________________________________________

*Finding the dominant species by leaf mass*
```{r}
dominant_leaf_species <- leaf_data %>%
  group_by(genus, species) %>%      
  summarise(total_leaf_mass_g = sum(mass_g, na.rm = TRUE)) %>% 
  arrange(desc(total_leaf_mass_g)) %>% 
  ungroup()

print(dominant_leaf_species)
```

Filter by plots
```{r}
N115_leaf_data <- leaf_data %>%
  filter(plot == "N-115")

HH115_leaf_data <- leaf_data %>%
  filter(plot == "HH-115")

B127_leaf_data <- leaf_data %>%
  filter(plot == "B-127")

U134_leaf_data <- leaf_data %>%
  filter(plot == "U-134")
```


```{r}
N115_dominant_leaf_species <- N115_leaf_data %>%
  group_by(genus, species) %>%      
  summarise(total_leaf_mass_g = sum(mass_g, na.rm = TRUE)) %>% 
  arrange(desc(total_leaf_mass_g)) %>% 
  ungroup()

print(N115_dominant_leaf_species)
```

```{r}
HH115_dominant_leaf_species <- HH115_leaf_data %>%
  group_by(genus, species) %>%      
  summarise(total_leaf_mass_g = sum(mass_g, na.rm = TRUE)) %>% 
  arrange(desc(total_leaf_mass_g)) %>% 
  ungroup()

print(HH115_dominant_leaf_species)
```

```{r}
B127_dominant_leaf_species <- B127_leaf_data %>%
  group_by(genus, species) %>%      
  summarise(total_leaf_mass_g = sum(mass_g, na.rm = TRUE)) %>% 
  arrange(desc(total_leaf_mass_g)) %>% 
  ungroup()

print(B127_dominant_leaf_species)
```

```{r}
U134_dominant_leaf_species <- U134_leaf_data %>%
  group_by(genus, species) %>%      
  summarise(total_leaf_mass_g = sum(mass_g, na.rm = TRUE)) %>% 
  arrange(desc(total_leaf_mass_g)) %>% 
  ungroup()

print(U134_dominant_leaf_species)
```

___________________________________________________________________________________________________________

*sub-setting my species of interest*

Need to account for both genus and species so we create what we want to get out
```{r}
target_species_combo <- list(
  c(genus = "Quercus", species = "alba"),
  c(genus = "Quercus", species = "rubra"),
  c(genus = "Acer", species = "saccharum")
)
```

Create an empty dataframe to store data once filtered 
  Sorry Christy, I saw you hate tibbles...
```{r}
species_interest_data <- tibble()
```

Loop through each target combination of species and genus and filter data
```{r}
for (s_combo in target_species_combo) {
  filtered_subset <- leaf_data %>%
    filter(genus == s_combo["genus"] & species == s_combo["species"])
  species_interest_data <- bind_rows(species_interest_data, filtered_subset)
}
```

```{r}
summary(species_interest_data)
```

```{r}
ggplot(data=species_interest_data) +
  facet_grid(year~plot) +
  # facet_wrap(~tissue, scales="free_y") +
  # geom_boxplot(aes(x=as.factor(week), y=mass_g, color=plot)) +
  geom_point(aes(x=week, y=mass_g, color=plot)) +
  stat_summary(geom="line", aes(x=week, y=mass_g), fun="mean") +
  labs(x="week", y="mass (g)") +
  scale_fill_manual(values=ewPlotColors) +
  scale_color_manual(values=ewPlotColors) +
  theme_bw()
```


```{r}
ggplot(data=species_interest_data) +
  facet_grid(year~species) +
  # facet_wrap(~tissue, scales="free_y") +
  # geom_boxplot(aes(x=as.factor(week), y=mass_g, color=plot)) +
  geom_point(aes(x=week, y=mass_g, color=species)) +
  stat_summary(geom="line", aes(x=week, y=mass_g), fun="mean") +
  labs(x="week", y="mass (g)") +
  theme_bw()
```
Creating a scientific name column to combine genus and species (mightve made earlier steps easier but alas we are here)
```{r}
species_interest_data$sci_name <- paste(species_interest_data$genus, species_interest_data$species, sep = " ")
summary(species_interest_data$sci_name)
head(species_interest_data$sci_name)
```
Calculate total mass per year per species
```{r}
leafSpTotal <- aggregate(mass_g ~ year + plot + trap_ID + sci_name,
                           data = species_interest_data,
                           FUN = sum,
                           na.action = na.omit) # Add na.action to handle NA values if any
names(leafSpTotal)[names(leafSpTotal)=="mass_g"] <- "totalMass_year_sci_name"
summary(leafSpTotal)
hist(leafSpTotal$totalMass_year_sci_name)
hist(leafSpTotal$totalMass_year_sci_name[leafSpTotal$year < max(leafSpTotal$year)]) 
summary(leafSpTotal[leafSpTotal$year < max(leafSpTotal$year),])
```
Merge Totals and calculate proportion
```{r}
# Merging our totals into the weekly sums so we can get proportion
species_interest_data <- merge(species_interest_data, leafSpTotal,
                               by = c("year", "plot", "trap_ID", "sci_name"),
                               all.x = TRUE) 

# Clean rows where trap_ID might be NA after merge (shouldn't happen if all.x=TRUE and trap_ID is always present)
species_interest_data <- species_interest_data[!is.na(species_interest_data$trap_ID),]

# Calculate the proportion
species_interest_data$mass_prop <- species_interest_data$mass_g / species_interest_data$totalMass_year_sci_name
summary(species_interest_data)

species_interest_data[is.na(species_interest_data$totalMass_year_scientific_name),]
```

Visualize leaf fall proportion <-- i hate this and hope to make something that looks better but this is a first start
```{r}
ggplot(data = species_interest_data) +
  facet_grid(year ~ plot + sci_name) +
  geom_point(aes(x = week, y = mass_prop, color = plot)) + 
  stat_summary(geom = "line", aes(x = week, y = mass_prop, group = interaction(plot, sci_name)), fun = "mean") +
  labs(x = "Week", y = "Proportion of Annual Leaf Fall") +
  scale_color_manual(values = ewPlotColors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))
```

Getting some summary stats
```{r}
aggLeafPlotSp <- aggregate(cbind(mass_g, mass_prop, totalMass_year_sci_name) ~ year + plot + week + date_collection + sci_name,
                         data = species_interest_data,
                         FUN = mean,
                         na.action = na.pass)
summary(aggLeafPlotSp)
```
Calculating peak weeks for the species level this time
```{r}
unique_years <- unique(species_interest_data$year)
unique_plots <- unique(species_interest_data$plot)
unique_sci_names <- unique(species_interest_data$sci_name)

weekPeak_sci_name <- data.frame(
  year = rep(unique_years, each = length(unique_plots) * length(unique_sci_names)),
  plot = rep(unique_plots, times = length(unique_years), each = length(unique_sci_names)),
  sci_name = rep(unique_sci_names, times = length(unique_years) * length(unique_plots)),
  week = NA,
  date = NA,
  propPeak = NA
)

# Remove the last year if it's incomplete (assuming it's max year)
weekPeak_sci_name <- weekPeak_sci_name[weekPeak_sci_name$year < max(weekPeak_sci_name$year),]

# Loop to populate weekPeak_scientific_name
for(YR in unique(weekPeak_sci_name$year)){
  datYr <- aggLeafPlotSp[aggLeafPlotSp$year==YR,] # Use aggLeafPlot from step 4
  for(PLT in unique(datYr$plot)){
    for(SCINAME in unique(datYr$sci_name)){ # Loop through scientific_names
      
      indNow <- which(weekPeak_sci_name$year==YR &
                      weekPeak_sci_name$plot==PLT &
                      weekPeak_sci_name$sci_name==SCINAME)
      datPlotSp <- datYr[datYr$plot==PLT & datYr$sci_name==SCINAME,]
      
          if(nrow(datPlotSp) > 0 && !all(is.na(datPlotSp$mass_prop))) {
        indPeak <- which.max(datPlotSp$mass_prop)
        
        if (length(indPeak) > 0) {
            wkPeak <- datPlotSp$week[indPeak]
            weekPeak_sci_name$week[indNow] <- wkPeak
            weekPeak_sci_name$date[indNow] <- as.character(datPlotSp$date_collection[indPeak])
            weekPeak_sci_name$propPeak[indNow] <- datPlotSp$mass_prop[indPeak]
        }
    }
  }
}
}
  
weekPeak_sci_name$date <- as.Date(weekPeak_sci_name$date)
summary(weekPeak_sci_name)
```
Again all of my graphs are looking really horrible, but I feel like I just have to get them out and if they are useful I will improve them. I need to figure out whats going on here bc it is hurting my brain
The jitter isn't doing what I want it to
```{r}
ggplot(data = weekPeak_sci_name, aes(x = year, y = week, color = plot, shape = sci_name)) +
  geom_point(size = 3, alpha = 1) +
   geom_jitter(width = 0.25, alpha = 0.5) +
  labs(x = "Year", y = "Week of Peak Leaf Fall") +
    scale_color_manual(values=ewPlotColors) +
  theme_bw() +
  scale_x_continuous(breaks = unique(weekPeak_sci_name$year))
```

HH115 QURU 2022 a bit odd 
```{r}
ggplot(data = weekPeak_sci_name, aes(x = year, y = week, color = plot, group = plot)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ sci_name, ncol = 1, scales = "free_y") +
  labs(x = "Year", y = "Week of Peak Leaf Fall") +
  scale_color_manual(values=ewPlotColors) +
  theme_bw() +
  scale_x_continuous(breaks = unique(weekPeak_sci_name$year)) +
  scale_y_continuous(breaks = seq(0, 52))
```

The old exchango chango showing the same thing but in a slightly different format, also lots of weird issues
```{r}
ggplot(data = weekPeak_sci_name,aes(x = year, y = week, color = sci_name, group = sci_name)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ plot, ncol = 2, scales = "free_y") +
  labs(x = "Year", y = "Week of Peak Leaf Fall") +
  theme_bw() +
  scale_x_continuous(breaks = unique(weekPeak_sci_name$year)) +
  scale_y_continuous(breaks = seq(0, 52))
```

```{r}
ggplot(data = aggLeafPlotSp) +
  facet_grid(plot ~ sci_name, scales = "free_y") + 
  geom_line(aes(x = week, y = mass_g, color = as.factor(year), group = as.factor(year))) +
  geom_point(aes(x = week, y = mass_g, color = as.factor(year))) +
  labs(x = "Week of Year", y = "Mass", color = "Year") +
  theme_bw() 
```
```{r}
ggplot(data = aggLeafPlotSp) +
  facet_grid(plot ~ sci_name, scales = "free_y") + 
  geom_line(aes(x = week, y = mass_prop, color = as.factor(year), group = as.factor(year))) +
  geom_point(aes(x = week, y = mass_prop, color = as.factor(year))) +
  labs(x = "Week of Year", y = "Proportion of Annual Leaf Fall", color = "Year") +
  theme_bw() 
```

```{r}
ggplot(data = aggLeafPlotSp[aggLeafPlotSp$week %in% 40:50 & aggLeafPlotSp$year %in% c(2018, 2021, 2022),]) +
  geom_line(aes(x = as.factor(week), y = mass_prop, color = as.factor(year)), size = 1, alpha = 0.7) +
  labs(x = "Week of Year", y = "Mass", color = "Year") +
  # coord_cartesian(xlim = c(40, 50)) +
  theme_bw() 
```

```{r}
ggplot(data = aggLeafPlotSp) +
  facet_wrap(~plot, ncol=2, scales="free_y") + 
  geom_line(aes(x = week, y = mass_g, color = as.factor(year), group = as.factor(year))) +
  geom_point(aes(x = week, y = mass_g, color = as.factor(year))) +
  labs(x = "Week of Year", y = "Proportion of Annual Leaf Fall", color = "Year") +
  theme_bw() 
```
```{r}
ggplot(data = aggLeafPlotSp) +
  facet_wrap(~sci_name, ncol=2, scales="free_y") + 
  geom_line(aes(x = week, y = mass_g, color = as.factor(year), group = as.factor(year))) +
  geom_point(aes(x = week, y = mass_g, color = as.factor(year))) +
  labs(x = "Week of Year", y = "Proportion of Annual Leaf Fall", color = "Year") +
  theme_bw() 
```
