---
title: "04_Leaflitter_timing"
author: "Cierra Lizer"
date: "2025-06-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

Set up file paths etc. --> this should also indicate where you can find these files!
```{r}
path.google <- "~/Google Drive/My Drive"
path.litter <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data")
path.figs <- file.path(path.litter, "figures") # where we shoudl save some figures
path.save <- file.path(path.litter, "LeafLitterData_Clean_forArchiving") # Where we shoudl save the data
```

Using a formatting theme consistent with what Meghan has done
```{r}
theme_base <-   theme(panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      panel.border = element_rect(fill="white", colour = "black", linewidth=0.7),
                      axis.title.x = element_text(margin = margin(t = 10, b=5), size=14),
                      axis.title.y = element_text(margin = margin(l = 5, r=5), size=14),
                      axis.text.x= element_text(margin = margin(t = 10), size=12),
                      axis.text.y=element_text(margin = margin(r = 10), size=12),
                      axis.ticks.length=unit(-0.3, "cm"),
                      # axis.ticks.margin=unit(0.5, "cm"),
                      axis.ticks = element_line(colour = "black", linewidth = 0.4))


# Setting a consistent color scheme across all graphs
plotOrder <- c("B-127", "U-134", "N-115", "HH-115")
ewPlotColors <- c("#1B9E77","#D95F02", "#7570B3", "#E7298A")
names(ewPlotColors) = plotOrder
ewPlotColors
```

**FIX THIS TO GO THROUGH FILE PATHS**
```{r}
dir(path.save)
all_cleaned_files <- list.files(path.save, pattern = "\\.csv$", full.names = TRUE)
if (length(all_cleaned_files) == 0) { stop(...) }
datLeafLitter <- map_df(all_cleaned_files, read_csv)
datLeafLitter$plot <- factor(datLeafLitter$plot, levels = plotOrder)
```

```{r}
datLeafLitter$year <- lubridate::year(datLeafLitter$date_collection)
datLeafLitter$yday <- lubridate::yday(datLeafLitter$date_collection)
datLeafLitter$week <- lubridate::week(datLeafLitter$date_collection)
```

Making a filter to have just the leaf tissue type
```{r}
leaf_data <- datLeafLitter %>%
  filter(tissue == "leaf")
```

Need to account for both genus and species so we create what we want to get out
```{r}
target_species_combo <- list(
  c(genus = "Quercus", species = "alba"),
  c(genus = "Quercus", species = "rubra"),
  c(genus = "Acer", species = "saccharum")
)
```

Create an empty dataframe to store data once filtered 
  Sorry Christy, I saw you hate tibbles...
```{r}
species_interest_data <- tibble()
```

Loop through each target combination of species and genus and filter data
```{r}
for (s_combo in target_species_combo) {
  filtered_subset <- leaf_data %>%
    filter(genus == s_combo["genus"] & species == s_combo["species"])
  species_interest_data <- bind_rows(species_interest_data, filtered_subset)
}
```

```{r}
summary(species_interest_data)
```

```{r}
species_interest_data$sci_name <- paste(species_interest_data$genus, species_interest_data$species, sep = " ")
summary(species_interest_data$sci_name)
head(species_interest_data$sci_name)
```

```{r}
leafSpTotal <- aggregate(mass_g ~ year + plot + trap_ID + sci_name,
                           data = species_interest_data,
                           FUN = sum,
                           na.action = na.omit) # Add na.action to handle NA values if any
names(leafSpTotal)[names(leafSpTotal)=="mass_g"] <- "totalMass_year_sci_name"
summary(leafSpTotal)
hist(leafSpTotal$totalMass_year_sci_name)
hist(leafSpTotal$totalMass_year_sci_name[leafSpTotal$year < max(leafSpTotal$year)]) 
summary(leafSpTotal[leafSpTotal$year < max(leafSpTotal$year),])
```
Merge Totals and calculate proportion
```{r}
# Merging our totals into the weekly sums so we can get proportion
species_interest_data <- merge(species_interest_data, leafSpTotal,
                               by = c("year", "plot", "trap_ID", "sci_name"),
                               all.x = TRUE) 

# Clean rows where trap_ID might be NA after merge (shouldn't happen if all.x=TRUE and trap_ID is always present)
species_interest_data <- species_interest_data[!is.na(species_interest_data$trap_ID),]

# Calculate the proportion
species_interest_data$mass_prop <- species_interest_data$mass_g / species_interest_data$totalMass_year_sci_name
summary(species_interest_data)

species_interest_data[is.na(species_interest_data$totalMass_year_scientific_name),]
```

```{r}
aggLeafPlotSp <- aggregate(cbind(mass_g, mass_prop, totalMass_year_sci_name) ~ year + plot + week + date_collection + sci_name,
                         data = species_interest_data,
                         FUN = mean,
                         na.action = na.pass)
summary(aggLeafPlotSp)
```


___________________________________________________________________________________________________________

*Differences in Plots*

Boxplots
```{r}
ggplot(data = aggLeafPlotSp[aggLeafPlotSp$week %in% 40:50 & aggLeafPlotSp$year %in% c(2018, 2021, 2022),]) +
  geom_boxplot(aes(x = as.factor(week), y = mass_g, color = as.factor(year))) +
  labs(x = "Week of Year", y = "Mass", color = "Year") +
  scale_color_manual(values=c("green4","blue3","purple2")) +
  # coord_cartesian(xlim = c(40, 50)) +
  theme_bw() 
```
Scatter of Medians
```{r}
ggplot(data = aggLeafPlotSp[aggLeafPlotSp$week %in% 40:50 & aggLeafPlotSp$year %in% c(2018, 2021, 2022),]) +
  stat_summary(fun = median, geom = "point",
         aes(x = as.factor(week), y = mass_g, color = as.factor(year), group = as.factor(year)), size=2) +
   geom_smooth(aes(x = as.factor(week), y = mass_g, color = as.factor(year), group = as.factor(year)),
              method = "loess", se = FALSE) + 
  labs(x = "Week of Year", y = "Mass", color = "Year") +
   scale_color_manual(values=c("green4","blue3","purple2")) +
  # coord_cartesian(xlim = c(40, 50)) +
  theme_bw()
```

ANOVA
```{r}
year_model <- aov(week ~ as.factor(year), data =aggLeafPlotSp)
summary(year_model)
```
```{r}
TukeyHSD(year_model)
```


___________________________________________________________________________________________________________

*Differences in Plots*

Boxplots
```{r}
ggplot(data = aggLeafPlotSp[aggLeafPlotSp$week %in% 40:50,]) +
  geom_boxplot(aes(x = as.factor(week), y = mass_g, color = as.factor(plot))) +
  labs(x = "Week of Year", y = "Mass", color = "plot") +
  # coord_cartesian(xlim = c(40, 50)) +
  scale_color_manual(values=ewPlotColors) +
  theme_bw() 
```

Scatter of Medians
```{r}
ggplot(data = aggLeafPlotSp[aggLeafPlotSp$week %in% 40:50,]) +
  stat_summary(fun = median, geom = "point",
               aes(x = as.factor(week), y = mass_g, color = plot, group = plot), size=2) +
   geom_smooth(aes(x = as.factor(week), y = mass_g, color = plot, group = plot),
              method = "loess", se = FALSE) + 
  labs(x = "Week of Year", y = "Mass", color = "Plot") +
  # coord_cartesian(xlim = c(40, 50)) +
  scale_color_manual(values=ewPlotColors) +
  theme_bw()
```

ANOVA
```{r}
plot_model <- aov(week ~ plot, data =aggLeafPlotSp)
summary(plot_model)
```
___________________________________________________________________________________________________________

*Differences in species*

Boxplots
```{r}
ggplot(data = aggLeafPlotSp[aggLeafPlotSp$week %in% 40:50,]) +
  geom_boxplot(aes(x = as.factor(week), y = mass_g, color = sci_name)) +
  labs(x = "Week of Year", y = "Mass", color = "Species") +
  # coord_cartesian(xlim = c(40, 50))
  scale_color_manual(values=c("red3","orange2","gold2")) +
  theme_bw() 
```

Scatter of Medians
```{r}
ggplot(data = aggLeafPlotSp[aggLeafPlotSp$week %in% 40:50,]) +
  stat_summary(fun = median, geom = "point",
               aes(x = as.factor(week), y = mass_g, color = sci_name, group = sci_name), size=2) +
   geom_smooth(aes(x = as.factor(week), y = mass_g, color = sci_name, group = sci_name),
              method = "loess", se = FALSE) + 
  labs(x = "Week of Year", y = "Mass", color = "Species") +
  scale_color_manual(values=c("red3","orange2","gold2")) +
  # coord_cartesian(xlim = c(40, 50)) +
  theme_bw()
```

Calculate total mass for each species per week
```{r}
total_mass_per_week_sp <- aggLeafPlotSp %>%
  group_by(sci_name, week) %>%
  summarise(total_mass_spec_week = sum(mass_g, na.rm = TRUE), .groups = 'drop_last') # .groups = 'drop_last' keeps sci_name grouped
```
From weekly total, calculate week with maximum mass for each species
```{r}
max_mass_week_species <- total_mass_per_week_sp %>%
  filter(total_mass_spec_week == max(total_mass_spec_week, na.rm = TRUE)) %>%
  ungroup()
print(max_mass_week_species)
```

ANOVA
```{r}
spec_model <- aov(week ~ sci_name, data =aggLeafPlotSp)
summary(spec_model)
```