---
title: "05_Leaflitter_nitrogen"
author: "Cierra Lizer"
date: "2025-07-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(lubridate)
library(dplyr)
library(readxl)
library(stringr)
library(readr)
```

Set up file paths etc. --> this should also indicate where you can find these files!
```{r}
path.google <- "~/Google Drive/My Drive"
path.litter <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data")
path.CN <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_runs")
path.figs <- file.path(path.litter, "figures") # where we shoudl save some figures
path.save <- file.path(path.litter, "LeafLitterData_Clean_forArchiving") # Where we shoudl save the data
```

**Need to switch this so it isn't just for me**
```{r}
cierra.path <- "G:/.shortcut-targets-by-id/0B_Fbr697pd36TkVHdDNJQ1dJU1E/East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_Runs"
```


Using a formatting theme consistent with what Meghan has done
```{r}
theme_base <-   theme(panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      panel.border = element_rect(fill="white", colour = "black", linewidth=0.7),
                      axis.title.x = element_text(margin = margin(t = 10, b=5), size=14),
                      axis.title.y = element_text(margin = margin(l = 5, r=5), size=14),
                      axis.text.x= element_text(margin = margin(t = 10), size=12),
                      axis.text.y=element_text(margin = margin(r = 10), size=12),
                      axis.ticks.length=unit(-0.3, "cm"),
                      # axis.ticks.margin=unit(0.5, "cm"),
                      axis.ticks = element_line(colour = "black", linewidth = 0.4))


# Setting a consistent color scheme across all graphs
plotOrder <- c("B-127", "U-134", "N-115", "HH-115")
ewPlotColors <- c("#1B9E77","#D95F02", "#7570B3", "#E7298A")
names(ewPlotColors) = plotOrder
ewPlotColors
```


Loading in data from 2021-22
```{r}
LLN2122path<-list.files(cierra.path, pattern = "Lizer_EWLL_2021-2022_run.xlsx", full.names = TRUE)
LLN2122 <- read_excel(LLN2122path)
head(LLN2122)
```

Name (aka sample id) will be composed of plot (first 4-5 characters), underscore, date (next 8 characters), underscore, species id (last 4 characters). Date is in the form year (4 characters), month (2 characters), day (2 characters)

Loading in 2018 data
```{r}
LLN18path<-list.files(cierra.path, pattern = "Fall_2018_run1.csv", full.names = TRUE)
LLN18 <- read_csv(LLN18path)
head(LLN18)
```
Name is composed of plot (1st character), dash, year (2 characters), month (2 character), day (2 characters), species id (2 characters)

Rename columns to be easier to read and pull information out from the name
```{r}
LLN2122_clean <- LLN2122 %>%
  rename(
    CN_ratio = `C/N  ratio`,
    N_percent = `N  [%]`,
    C_percent = `C  [%]`
  ) %>%
  
  mutate(
    spec_id = str_sub(Name, -4), # Get the last 4 characters
    date_raw = str_sub(Name, -13, -5), #finds the 8 characters before the last underscore (5th to last character)
    #note: will give year, month, day in a string together
    plot = str_extract(Name, "^[^_]+") # Extracts characters from start (^) until the first underscore (_)
  ) %>%
#convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, everything(), -date_raw # Keep original, then new columns, then rest
  )

print(LLN2122_clean)
```

A BUNCH of cleaning 2018 and making the formats similar to what I had done
```{r}
LLN18_clean <- LLN18 %>%
  rename(
    CN_ratio = `C:N`,
    N_percent = `%N`,
    C_percent = `%C`
  ) %>%
  mutate(
    spec_id_raw = str_sub(Name, 11,12),
    year_2dig = str_sub(Name, 3, 4),
    month_2dig = str_sub(Name, 6,7),
    day_2dig = str_sub(Name, 9,10),
    plot_raw = str_sub(Name, 1,1)
  ) %>%
  mutate(
    spec_id = case_when(
      spec_id_raw == "QA" ~ "QUAL",
      spec_id_raw == "TA" ~ "TIAM",
      spec_id_raw == "AS" ~ "ACSA",
      TRUE ~ spec_id_raw
    ), 
    plot = case_when(
      plot_raw == "H" ~ "HH115",
      plot_raw == "N" ~ "N115",
      plot_raw == "U" ~ "U134",
      plot_raw == "B" ~ "B127",
      TRUE ~ plot_raw
    )
  ) %>% 
  mutate(
    date_raw = paste0("20", year_2dig, month_2dig, day_2dig)
  ) %>%
  # convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, CN_ratio, N_percent, C_percent, everything(),
    -year_2dig, -month_2dig, -day_2dig, -date_raw, -plot_raw, -spec_id_raw
  )

print(LLN18_clean)
unique(LLN18_clean$spec_id)
unique(LLN18_clean$plot)
```

AGGGGHHH I accidentally put QUAB and QUAL both meaning Quercus alba
```{r}
LLN2122_clean <- LLN2122_clean %>%
  mutate(
    spec_id = case_when(
    spec_id == "QUAB" ~ "QUAL", # If spec_id is "QUAB", change it to "QUAL"
    TRUE ~ spec_id
  )
  )


LLN2122_clean <- LLN2122_clean %>%
  mutate(
    plot = case_when(
    plot == "H115" ~ "HH115",
    TRUE ~ plot
    )
  )

unique(LLN2122_clean$spec_id)
unique(LLN2122_clean$plot)
```

Combining my 2 dataframes
```{r}
LLNdattot <- bind_rows(LLN2122_clean, LLN18_clean)

dim(LLNdattot)
unique(LLNdattot$year)
```

Removing Tilia Americana from the dataset
```{r}
LLN_spec <- LLNdattot %>%
  filter(spec_id != "TIAM")

unique(LLN_spec$spec_id)
dim(LLN_spec)
```
```{r}
write.csv(LLN_spec, file.path(path.google, "URF REU 2025 - Lizer - Leaf Litter ", "CNcombined.csv"), row.names=F)
```



Yearly Differences
```{r}
ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = as.factor(year))) +
   facet_grid(~spec_id) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week", y = "N_percent", color = "Year") +
  theme_minimal()
```

Plot
```{r}
ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = plot)) + 
     facet_grid(~spec_id) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week", y = "C/N Ratio", color = "plot") +
  scale_color_manual(values=ewPlotColors) +
  theme_minimal()
```

Species
```{r}
ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = spec_id)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week", y = "C/N Ratio", color = "Species") +
  theme_minimal()
```


```{r}
lmeNyear <- lme(N_percent ~ as.factor(year), random=list(spec_id=~1, plot=~1, week=~1), data=LLN_spec)
summary(lmeNyear)
anova(lmeNyear)
```

```{r}
mean__by_year_species <- aggLeafSppTot %>%
  group_by(year, sci_name) %>%
  summarize(mean_weekPeak = mean(weekPeak, na.rm = TRUE))
print(mean_wkpk_by_year_species)
```

