---
title: "07_Leaflitter_plottrap"
author: "Cierra Lizer"
date: "2025-07-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(lubridate)
library(dplyr)
library(readxl)
library(stringr)
library(readr)
library(nlme);  # Does the mixed effects model
library(emmeans) # will et us do a multi-comparisons test
```

```{r}
theme_base <-   theme(panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      panel.border = element_rect(fill="white", colour = "black", linewidth=0.7),
                      axis.title.x = element_text(margin = margin(t = 10, b=5), size=14),
                      axis.title.y = element_text(margin = margin(l = 5, r=5), size=14),
                      axis.text.x= element_text(margin = margin(t = 10), size=12),
                      axis.text.y=element_text(margin = margin(r = 10), size=12),
                      axis.ticks.length=unit(-0.3, "cm"),
                      # axis.ticks.margin=unit(0.5, "cm"),
                      axis.ticks = element_line(colour = "black", linewidth = 0.4))


# Setting a consistent color scheme across all graphs
plotOrder <- c("B-127", "U-134", "N-115", "HH-115")
ewPlotColors <- c("#1B9E77","#D95F02", "#7570B3", "#E7298A")
names(ewPlotColors) = plotOrder
ewPlotColors
```

```{r}
path.google <- "~/Google Drive/My Drive"
path.litter <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data")
path.CN <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_runs")
path.figs <- file.path(path.litter, "figures") # where we shoudl save some figures
path.save <- file.path(path.litter, "LeafLitterData_Clean_forArchiving") # Where we shoudl save the data
cierra.path <- "G:/.shortcut-targets-by-id/0B_Fbr697pd36TkVHdDNJQ1dJU1E/East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_Runs"
```

___________________________________________________________________________________________________________________________________________

*Cleaning Leaf litter*

```{r}
#combine into one dataset
all_cleaned_files <- list.files(path.save, pattern = "\\.csv$", full.names = TRUE)
if (length(all_cleaned_files) == 0) { stop(...) }
datLitter <- map_df(all_cleaned_files, read_csv)
datLitter$plot <- factor(datLitter$plot, levels = plotOrder)
summary(datLitter)

datLitter$year <- lubridate::year(datLitter$date_collection)
datLitter$yday <- lubridate::yday(datLitter$date_collection)
datLitter$week <- lubridate::week(datLitter$date_collection)
summary(datLitter)

datLitter <- datLitter[datLitter$year<2024,]
summary(datLitter)
```

Cleaning of missing and empty's
```{r}
# There is no HH-115 NE, NW, SE, SW --> it has a weird layout
#datLitter[datLitter$plot=="HH-115" & datLitter$trap_ID %in% c("NE", "NW", "SE", "SW"), "trap_ID"] <- NA

datLitter <- datLitter[!is.na(datLitter$trap_ID),]

# Finding bags that that are empty or missing
unique(datLitter$tissue)

datMissing <- datLitter[datLitter$tissue == "MISSING DATA",]
dim(datMissing)
summary(datMissing)

datNone <- datLitter[datLitter$tissue == "EMPTY BAG",]
dim(datNone)
summary(datNone)
```

```{r}
datLitter_cleaned <- datLitter %>%
  mutate(sci_name = paste(genus, species, sep = " ")) %>%
  filter(!is.na(plot)) %>% # Filter out NA plots
  filter(!is.na(trap_ID)) %>% # Filter out NA trap_IDs
  filter(tissue == "leaf") # Filter to leaf tissue only
```


Summing to the trap level --> good to use for comparisons of plots and forest wide things
```{r}
aggLeafTrap <- datLitter_cleaned %>%
  group_by(year, week, date_collection, plot, trap_ID) %>%
  summarise(mass_g = sum(mass_g, na.rm = TRUE), .groups = 'drop')
```

```{r}
allCollect = NULL
# plotsCore <- c("N", "S", "E", "W")
# plotExtra <- c("NW", "SW", "NE", "SE")
for(PLT in unique(aggLeafTrap$plot)){
  datPLT <- aggLeafTrap[aggLeafTrap$plot==PLT,]
  datPLT <- droplevels(datPLT)
  
  datePLT <- unique(datPLT$date_collection)
  datePLT <- sort(datePLT)
  daysPLT <- diff(datePLT)
  # Making a data frame with all collections dates and traps for the plot
  # That way we can add in plots as missing if they're not there
  dfCollection <- data.frame(plot=PLT, trap_ID = rep(unique(datPLT$trap_ID), each=length(datePLT)),
                             date_collection=datePLT, days_collection=c(NA, daysPLT))
                            
  
  if(is.null(allCollect)){
    allCollect <- dfCollection
  } else {
    allCollect <- rbind(allCollect, dfCollection)
  }
}
summary(allCollect)

# Merging the number of days in each collection period with the actual data

dim(aggLeafTrap)
aggLeafTrap <- merge(aggLeafTrap, allCollect, all=T)
aggLeafTrap$year <- lubridate::year(aggLeafTrap$date_collection)
aggLeafTrap$week <- lubridate::week(aggLeafTrap$date_collection)
summary(aggLeafTrap)



# This added in a lot of NAs because we added half our traps a couple years in
for(PLT in unique(aggLeafTrap$plot)){
  for(TRP in unique(aggLeafTrap$trap_ID[aggLeafTrap$plot==PLT])){
    minTrap <- min(aggLeafTrap$date_collection[aggLeafTrap$plot==PLT & aggLeafTrap$trap_ID==TRP & !is.na(aggLeafTrap$mass_g)] ,na.rm=T)
    
    # test <- aggLeafTrap[aggLeafTrap$plot==PLT & aggLeafTrap==TRP & aggLeafTrap$date_collection>minTrap,]
    
    aggLeafTrap <- aggLeafTrap[aggLeafTrap$plot!=PLT | (aggLeafTrap$plot==PLT & aggLeafTrap$trap_ID!=TRP) | 
                                 (aggLeafTrap$plot==PLT & aggLeafTrap$trap_ID==TRP & aggLeafTrap$date_collection>minTrap),]
  }
}


dim(aggLeafTrap)
summary(aggLeafTrap)

aggLeafTrap$mass_g_day <- aggLeafTrap$mass_g/aggLeafTrap$days_collection
summary(aggLeafTrap)
```

```{r}
# Lets calculate the proportion of leaf fall at any given point in time
leafTrapTotal <- aggregate(mass_g ~ year + plot + trap_ID, data=aggLeafTrap, FUN=sum)
names(leafTrapTotal)[names(leafTrapTotal)=="mass_g"] <- "totalMass_year"
summary(leafTrapTotal)
summary(leafTrapTotal[leafTrapTotal$year<2024,])

# Merging our totals into the weekly sums so we can get proportion
aggLeafTrap <- merge(aggLeafTrap, leafTrapTotal, all=T)
aggLeafTrap <- aggLeafTrap[!is.na(aggLeafTrap$trap_ID),]
aggLeafTrap$mass_prop <- aggLeafTrap$mass_g/aggLeafTrap$totalMass_year
summary(aggLeafTrap)
aggLeafTrap[is.na(aggLeafTrap$totalMass_year),]
```

Plot level --> Will need for nitrogen analysis
```{r}
# Getting some plot-level summary stats
aggLeafPlot <- aggLeafTrap %>%
  group_by(year, plot, week, date_collection) %>%
  summarise(
    mass_g = mean(mass_g, na.rm = TRUE), # Mean mass_g across traps for this plot-week
    mass_g_day = mean(mass_g_day, na.rm = TRUE), # Mean mass_g_day across traps
    mass_prop = mean(mass_prop, na.rm = TRUE), # Mean mass_prop across traps
    totalMass_year = first(totalMass_year), # totalMass_year should be consistent per plot-year
    .groups = 'drop'
  )
summary(aggLeafPlot)

aggLeafPlot_by_species <- datLitter_cleaned %>%
  group_by(year, week, date_collection, plot, sci_name) %>%
  summarise(mass_g_species = sum(mass_g, na.rm = TRUE), .groups = 'drop')
summary(aggLeafPlot_by_species)
```

```{r}
weekPeak <- data.frame(year=rep(unique(aggLeafTrap$year), each=length(unique(aggLeafPlot$plot))),
                       plot=rep(unique(aggLeafPlot$plot)),
                       week=NA,
                       date=NA,
                       propPeak = NA,
                       prop9Wk = NA,
                       prop5Wk = NA,
                       prop3Wk = NA)

for(YR in unique(weekPeak$year)){
  datYr <- aggLeafPlot[aggLeafPlot$year==YR,]
  for(PLT in unique(datYr$plot)){
    indNow <- which(weekPeak$year==YR & weekPeak$plot==PLT)
    datPlot <- datYr[datYr$plot==PLT,]
    
    indPeak <- which(datPlot$mass_g_day==max(datPlot$mass_g_day))
    wkPeak <- datPlot$week[indPeak]
    weekPeak$week[indNow] <- wkPeak
    weekPeak$date[indNow] <- as.character(datPlot$date_collection[indPeak])
    weekPeak$propPeak[indNow] <- datPlot$mass_prop[indPeak]
    weekPeak$prop9Wk[indNow] <- sum(datPlot$mass_prop[datPlot$week %in% (wkPeak-4):(wkPeak + 4)])
    weekPeak$prop5Wk[indNow] <- sum(datPlot$mass_prop[datPlot$week %in% (wkPeak-2):(wkPeak + 2)])
    weekPeak$prop3Wk[indNow] <- sum(datPlot$mass_prop[datPlot$week %in% (wkPeak-1):(wkPeak + 1)])
  }
}
weekPeak$date <- as.Date(weekPeak$date)
summary(weekPeak)
weekPeak[weekPeak$prop5Wk<0.5,]

#write.csv(weekPeak, file.path(path.google, "URF REU 2025 - Lizer - Leaf Litter ", "PeakLeafDates_byPlot.csv"), row.names=F)
```

```{r}
summary(leafTrapTotal)
for(YR in unique(aggLeafTrap$year)){
  print(YR)
  datYr <- aggLeafTrap[aggLeafTrap$year==YR,]
  for(PLT in unique(datYr$plot)){
    print(PLT)
    datPlot <- datYr[datYr$plot==PLT,]
    for(TRP in unique(datPlot$trap_ID)){
      indNow <- which(leafTrapTotal$year==YR & leafTrapTotal$plot==PLT & leafTrapTotal$trap_ID==TRP)
      datTrp <- datPlot[datPlot$trap_ID==TRP,]
  
      print(paste(TRP, "-", indNow))
      indPeak <- which(datTrp$mass_g_day==max(datTrp$mass_g_day, na.rm=T))
      wkPeak <- datTrp$week[indPeak]
      # tmpPleafTrapTotaleak <- data.frame(year = YR, plot=PLT, trap_ID=TRP, week=NA, date=NA, propPeak=NA)
      leafTrapTotal[indNow,"weekPeak"] <- wkPeak
      leafTrapTotal[indNow,"date"] <- as.character(datTrp$date_collection[indPeak])
      leafTrapTotal[indNow,"propPeak"] <- datTrp$mass_prop[indPeak]
      # tmpPeak[1,"prop9Wk"] <- sum(datTrp$mass_prop[datTrp$week %in% (wkPeak-4):(wkPeak + 4)])
      # tmpPeak[1,"prop5Wk"] <- sum(datTrp$mass_prop[datTrp$week %in% (wkPeak-2):(wkPeak + 2)])
      # tmpPeak[1,"prop3Wk"] <- sum(datTrp$mass_prop[datTrp$week %in% (wkPeak-1):(wkPeak + 1)])
    }
  }
}
# weekPeakTrap$date <- as.Date(weekPeakTrap$date)
summary(leafTrapTotal)
summary(leafTrapTotal[is.na(leafTrapTotal$weekPeak),])
# weekPeak[weekPeak$prop5Wk<0.5,]
```

___________________________________________________________________________________________________________________________________________

*LL & Met*
Want it aggregated at the trap level so we can have replicates within plots

```{r}
metSummer <- read.csv("G:/.shortcut-targets-by-id/1q2wvODXrDo0tgOTLpFqF7TqcWoKoHZjW/URF REU 2025 - Lizer - Leaf Litter/data/daymet/daymet_June-July-August_summaries_2017-2023.csv")
summary(metSummer)
```

```{r}
leafTrapMet <- merge(leafTrapTotal, metSummer, all.x=T)
summary(leafTrapMet)
```

Filter to be just the years I did nitrogen analysis on
```{r}
years_for_nitrogen_analysis <- c(2018, 2020, 2021, 2022)

leafTrapMet_filter <- leafTrapMet %>%
  filter(year %in% years_for_nitrogen_analysis)
```


```{r}
lmeYrs <- lme(weekPeak ~ as.factor(year), random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeYrs)$tTable
anova(lmeYrs)

lmeLLRainless <- lme(weekPeak ~ n.Rainless, random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLRainless)
anova(lmeLLRainless)

lmeLLRainlessConsec <- lme(weekPeak ~ RainlessConsec.max , random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLRainlessConsec)
anova(lmeLLRainlessConsec)

lmeLLPrcp <- lme(weekPeak ~ prcp..mm.day., random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLPrcp)
anova(lmeLLPrcp)

lmeLLTmax <- lme(weekPeak ~ tmax..deg.c., random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLTmax)$tTable
anova(lmeLLTmax)

lmeLLVP <- lme(weekPeak ~ vp..Pa., random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLVP)
anova(lmeLLVP)

lmeLLVPD <- lme(weekPeak ~ VPD.avg, random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLVPD)$tTable
anova(lmeLLVPD)
```

```{r}
ggplot(data=leafTrapMet) +
  geom_boxplot(aes(x=as.factor(year), y=weekPeak)) +
  geom_point(aes(x=as.factor(year), y=weekPeak, color=plot), position=position_jitter(0.1))


ggplot(data = leafTrapMet_filter, aes(x = n.Rainless, y = weekPeak)) +
  geom_point(aes(color=as.character(year)), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Number of Rainless Days", y = "Week Peak") +
  theme_minimal()

ggplot(data=leafTrapMet_filter, aes(x=prcp..mm.day., y=weekPeak)) +
  # geom_boxplot(aes(x=as.factor(year), y=weekPeak)) +
  geom_point(aes(color=plot), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)


ggplot(data=leafTrapMet_filter, aes(x=RainlessConsec.max, y=weekPeak))+
  geom_point(aes(color=plot), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)

ggplot(data=leafTrapMet_filter, aes(x=tmax..deg.c., y=weekPeak))+
  geom_point(aes(color=plot), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)

ggplot(data=leafTrapMet_filter, aes(x=vp..Pa., y=weekPeak))+
  geom_point(aes(color=plot), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)

ggplot(data=leafTrapMet_filter, aes(x=VPD.avg, y=weekPeak))+
  geom_point(aes(color=plot), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

___________________________________________________________________________________________________________________________________________

*Cleaning N data*
Loading in data from 2021-22
```{r}
LLN2122path<-list.files(cierra.path, pattern = "Lizer_EWLL_2021-2022_run.xlsx", full.names = TRUE)
LLN2122 <- read_excel(LLN2122path)
head(LLN2122)
```

Name (aka sample id) will be composed of plot (first 4-5 characters), underscore, date (next 8 characters), underscore, species id (last 4 characters). Date is in the form year (4 characters), month (2 characters), day (2 characters)

Loading in 2018 data
```{r}
LLN18path<-list.files(cierra.path, pattern = "Fall_2018_run1.csv", full.names = TRUE)
LLN18 <- read_csv(LLN18path)
head(LLN18)
```
Name is composed of plot (1st character), dash, year (2 characters), month (2 character), day (2 characters), species id (2 characters)

Load in Cierra's 2nd run (has some 2018, 2020, and 2021 data)
```{r}
LLNrun2path<-list.files(cierra.path, pattern = "Lizer_EWLL_run2_0715_corrected.xlsx", full.names = TRUE)
LLNrun2 <- read_excel(LLNrun2path)
head(LLNrun2)
```

Load in Cierra's 3rd run (mostly 2020 and a few 2018)
```{r}
LLNrun3path<-list.files(cierra.path, pattern = "Lizer_EWLL_Run3_SH_072125.xlsx", full.names = TRUE)
LLNrun3 <- read_excel(LLNrun3path)
head(LLNrun3)
```

Rename columns to be easier to read and pull information out from the name
```{r}
LLN2122_clean <- LLN2122 %>%
  rename(
    CN_ratio = `C/N  ratio`,
    N_percent = `N  [%]`,
    C_percent = `C  [%]`
  ) %>%
  
  mutate(
    spec_id = str_sub(Name, -4), # Get the last 4 characters
    date_raw = str_sub(Name, -13, -5), #finds the 8 characters before the last underscore (5th to last character)
    #note: will give year, month, day in a string together
    plot = str_extract(Name, "^[^_]+") # Extracts characters from start (^) until the first underscore (_)
  ) %>%
#convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, everything(), -date_raw # Keep original, then new columns, then rest
  )

print(LLN2122_clean)
```

A BUNCH of cleaning 2018 and making the formats similar to what I had done
```{r}
LLN18_clean <- LLN18 %>%
  rename(
    CN_ratio = `C:N`,
    N_percent = `%N`,
    C_percent = `%C`
  ) %>%
  mutate(
    spec_id_raw = str_sub(Name, 11,12),
    year_2dig = str_sub(Name, 3, 4),
    month_2dig = str_sub(Name, 6,7),
    day_2dig = str_sub(Name, 9,10),
    plot_raw = str_sub(Name, 1,1)
  ) %>%
  mutate(
    spec_id = case_when(
      spec_id_raw == "QA" ~ "QUAL",
      spec_id_raw == "TA" ~ "TIAM",
      spec_id_raw == "AS" ~ "ACSA",
      TRUE ~ spec_id_raw
    ), 
    plot = case_when(
      plot_raw == "H" ~ "HH115",
      plot_raw == "N" ~ "N115",
      plot_raw == "U" ~ "U134",
      plot_raw == "B" ~ "B127",
      TRUE ~ plot_raw
    )
  ) %>% 
  mutate(
    date_raw = paste0("20", year_2dig, month_2dig, day_2dig)
  ) %>%
  # convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, CN_ratio, N_percent, C_percent, everything(),
    -year_2dig, -month_2dig, -day_2dig, -date_raw, -plot_raw, -spec_id_raw
  )

print(LLN18_clean)
unique(LLN18_clean$spec_id)
unique(LLN18_clean$plot)
```

Cleaning of Run 2
```{r}
LLNrun2_select <- LLNrun2 %>%
  select(Name, Method, `N  [%]`, `C  [%]`, `C/N  ratio`)

LLNrun2_clean <- LLNrun2_select %>%
  mutate(
    `N  [%]` = as.numeric(`N  [%]`), 
    `C  [%]` = as.numeric(`C  [%]`),
    `C/N  ratio` = as.numeric(`C/N  ratio`)
  )%>%

#need to take out blanks and standards, some are labeled with different methods, some have values of 0.
 filter(Method == "Plant") %>%
 filter(!(`N  [%]` == 0 & `C  [%]` == 0 & `C/N  ratio` == 0)) %>%
  
  rename(
    CN_ratio = `C/N  ratio`,
    N_percent = `N  [%]`,
    C_percent = `C  [%]`
  ) %>%
  
  mutate(
    spec_id = str_sub(Name, -4), # Get the last 4 characters
    date_raw = str_sub(Name, -13, -5), #finds the 8 characters before the last underscore (5th to last character)
    #note: will give year, month, day in a string together
    plot = str_extract(Name, "^[^_]+") # Extracts characters from start (^) until the first underscore (_)
  ) %>%
#convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, everything(), -date_raw, -Method # Keep original, then new columns, then rest
  )

LLNrun2_clean <- LLNrun2_clean %>%
  mutate(
    plot = case_when(
    plot == "H115" ~ "HH115",
    plot == "HH15" ~ "HH115",
    TRUE ~ plot
    )
  )

print(LLNrun2_clean)
unique(LLNrun2_clean$spec_id)
unique(LLNrun2_clean$plot)
```


Cleaning of Run 3
```{r}
LLNrun3_select <- LLNrun3 %>%
  select(Name, Method, `N  [%]`, `C  [%]`, `C/N  ratio`)

LLNrun3_clean <- LLNrun3_select %>%
  mutate(
    `N  [%]` = as.numeric(`N  [%]`), 
    `C  [%]` = as.numeric(`C  [%]`),
    `C/N  ratio` = as.numeric(`C/N  ratio`)
  )%>%

#need to take out blanks and standards, some are labeled with different methods, some have values of 0.
 filter(Method == "Plant") %>%
 filter(!(`N  [%]` == 0 & `C  [%]` == 0 & `C/N  ratio` == 0)) %>%
  
  rename(
    CN_ratio = `C/N  ratio`,
    N_percent = `N  [%]`,
    C_percent = `C  [%]`
  ) %>%
  
  mutate(
    spec_id = str_sub(Name, -4), # Get the last 4 characters
    date_raw = str_sub(Name, -13, -5), #finds the 8 characters before the last underscore (5th to last character)
    #note: will give year, month, day in a string together
    plot = str_extract(Name, "^[^_]+") # Extracts characters from start (^) until the first underscore (_)
  ) %>%
#convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, everything(), -date_raw, -Method # Keep original, then new columns, then rest
  )

LLNrun3_clean <- LLNrun3_clean %>%
  mutate(
    plot = case_when(
    plot == "H115" ~ "HH115",
    plot == "HH15" ~ "HH115",
    TRUE ~ plot
    )
  )

print(LLNrun3_clean)
unique(LLNrun3_clean$spec_id)
unique(LLNrun3_clean$plot)
```

AGGGGHHH I accidentally put QUAB and QUAL both meaning Quercus alba
```{r}
LLN2122_clean <- LLN2122_clean %>%
  mutate(
    spec_id = case_when(
    spec_id == "QUAB" ~ "QUAL", # If spec_id is "QUAB", change it to "QUAL"
    TRUE ~ spec_id
  )
  )


LLN2122_clean <- LLN2122_clean %>%
  mutate(
    plot = case_when(
    plot == "H115" ~ "HH115",
    TRUE ~ plot
    )
  )

unique(LLN2122_clean$spec_id)
unique(LLN2122_clean$plot)
```

Combining my 3 dataframes
```{r}
LLNdattot <- bind_rows(LLN2122_clean, LLN18_clean, LLNrun2_clean, LLNrun3_clean)

LLNdattot <- LLNdattot %>%
  mutate(year = as.numeric(as.character(year)))

dim(LLNdattot)
unique(LLNdattot$year)
```

Removing Tilia Americana from the dataset
```{r}
LLN_spec <- LLNdattot %>%
  filter(spec_id != "TIAM")

unique(LLN_spec$spec_id)
dim(LLN_spec)
```

```{r}
#write.csv(LLN_spec, file.path(path.google, "URF REU 2025 - Lizer - Leaf Litter ", "CNcombined.csv"), row.names=F)
```

```{r}
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "QUAL" ~ "Quercus alba", 
    TRUE ~ spec_id
  )
  )
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "QURU" ~ "Quercus rubra", 
    TRUE ~ sci_name
  )
  )
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "ACSA" ~ "Acer saccharum", 
    TRUE ~ sci_name
  )
  )

head(LLN_spec)
unique(LLN_spec$sci_name)
```
___________________________________________________________________________________________________________________________________________
*Looking at just N*
```{r}
ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = plot)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week", y = "%N", color = "plot") +
  theme_minimal()

ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = plot)) + 
  facet_grid(~sci_name) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week", y = "%N", color = "plot") +
  theme_minimal()

ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = as.factor(year))) + 
  facet_grid(~sci_name) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week", y = "%N", color = "year") +
  theme_minimal()
```

```{r}
lmeNyear <- lme(N_percent ~ as.factor(year), random=list(spec_id=~1, plot=~1, week=~1), data=LLN_spec)
summary(lmeNyear)
anova(lmeNyear)
```

___________________________________________________________________________________________________________________________________________
*N and Met data*

```{r}
NMet <- merge(LLN_spec, metSummer,  by = c("year"), all.x=T)
summary(NMet)
```

```{r}
ggplot(data = NMet, aes(x = n.Rainless, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Number of Rainless Days", y = "%N", color = "Species") +
  theme_minimal()

ggplot(data = NMet, aes(x = prcp..mm.day., y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Number of Rainless Days", y = "%N", color = "Species") +
  theme_minimal()

ggplot(data = NMet, aes(x = RainlessConsec.max, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Number of Consecutive Rainless Days", y = "%N") +
  theme_minimal()

ggplot(data = NMet, aes(x = tmax..deg.c., y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Maximum Temperature", y = "%N") +
  theme_minimal()

ggplot(data = NMet, aes(x = vp..Pa., y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Vapor Pressure", y = "%N") +
  theme_minimal()
```

```{r}
lmeNRainless <- lme(N_percent ~ n.Rainless, random=list(plot=~1), data=NMet)
summary(lmeNRainless)
anova(lmeNRainless)

lmeNRainlessConsec <- lme(N_percent ~ RainlessConsec.max, random=list(plot=~1), data=NMet)
summary(lmeNRainlessConsec)
anova(lmeNRainlessConsec)

lmeNPrcp <- lme(N_percent ~ prcp..mm.day., random=list(plot=~1), data=NMet)
summary(lmeNPrcp)
anova(lmeNPrcp)

lmeNTmax <- lme(N_percent ~ tmax..deg.c., random=list(plot=~1), data=NMet)
summary(lmeNTmax)
anova(lmeNTmax)

lmeNVP <- lme(N_percent ~ vp..Pa., random=list(plot=~1), data=NMet)
summary(lmeNVP)
anova(lmeNVP)
```


___________________________________________________________________________________________________________________________________________

*N and LL data*
Making sure N and LL data align
```{r}
unique(LLN_spec$plot)
unique(leafTrapTotal$plot)

LLN_spec <- LLN_spec %>%
  mutate(plot = recode(plot,
  "B127" = "B-127",
  "N115" = "N-115",
  "HH115" = "HH-115",
  "U134" = "U-134"
  ))

unique(LLN_spec$plot)

```

```{r}
aggLeafPlot_by_species <- datLitter_cleaned %>%
  mutate(sci_name = paste(genus, species, sep = " ")) %>%
  group_by(year, week, date_collection, plot, sci_name) %>%
  summarise(mass_g_species = sum(mass_g, na.rm = TRUE), .groups = 'drop') %>%
  mutate(year = as.character(year)) 

total_plot_week_mass <- aggLeafPlot_by_species %>%
  group_by(year, plot, week, date_collection) %>%
  summarise(total_mass_g_for_peak_calc = sum(mass_g_species, na.rm = TRUE), .groups = 'drop') %>%
  # Join with allCollect to get `days_collection` for `mass_g_day` calculation at plot level
  left_join(allCollect %>% select(plot, date_collection, days_collection) %>% distinct(), # Use distinct to avoid duplicates
            by = c("plot", "date_collection")) %>%
  mutate(mass_g_day_for_peak_calc = total_mass_g_for_peak_calc / days_collection)
summary(total_plot_week_mass)

weekPeak_plot_level_summary <- total_plot_week_mass %>%
  group_by(year, plot) %>%
  # Filter for max mass_g_day (handling all NA cases)
  filter(if(all(is.na(mass_g_day_for_peak_calc))) TRUE else mass_g_day_for_peak_calc == max(mass_g_day_for_peak_calc, na.rm = TRUE)) %>%
  rename(weekPeak = week) %>%
  ungroup() %>%
  select(year, plot, weekPeak, date = date_collection) 

weekPeak_plot_level_summary <- total_plot_week_mass %>%
  group_by(year, plot) %>%
  filter(if(all(is.na(mass_g_day_for_peak_calc))) TRUE else mass_g_day_for_peak_calc == max(mass_g_day_for_peak_calc, na.rm = TRUE)) %>%
  rename(weekPeak = week) %>%
  ungroup() %>%
  # *** REMOVE 'date = date_collection' from this select statement ***
  select(year, plot, weekPeak)

totalMass_year_per_plot <- aggLeafPlot_by_species %>%
    group_by(year, plot) %>%
    summarise(totalMass_year_plot = sum(mass_g_species, na.rm = TRUE), .groups = 'drop')

aggwkpkPlot <- left_join(
  weekPeak_plot_level_summary, # This no longer has 'date'
  totalMass_year_per_plot,
  by = c("year", "plot")
) %>%
  mutate(year = as.character(year))

```


OG code without mass
```{r}
aggwkpkPlot <- leafTrapTotal %>%
  group_by(year, plot) %>%
  summarise(
    weekPeak = mean(weekPeak, na.rm = TRUE), 
    totalMass_year = first(totalMass_year),  
    .groups = 'drop'
  ) %>%
  mutate(year = as.character(year))
dim(leafTrapTotal)
dim(aggwkpkPlot)
head(aggwkpkPlot)

LLN_spec <- LLN_spec %>%
  mutate(year = as.character(year))

LLN_plot <- LLN_spec %>%
  left_join(aggwkpkPlot, by = c("year", "plot"))
##still something wrong with this
dim(LLN_plot)
dim(LLN_spec)
head(LLN_plot)

LLN_peak <- LLN_plot %>%
  filter(week >= round(weekPeak) - 1 & week <= round(weekPeak) + 1)
dim(LLN_peak)
head(LLN_peak)
```

new merge with masses to get total N
```{r}
#apparently there is a duplicate in LLN_spec
LLN_spec_unique <- LLN_spec %>%
  group_by(year, plot, week, sci_name) %>%
  summarise(N_percent = mean(N_percent, na.rm = TRUE), .groups = 'drop')
print(LLN_spec_unique %>% count(year, plot, week, sci_name) %>% filter(n > 1)) #should be zero


aggLeafPlot_by_species <- aggLeafPlot_by_species %>%
  mutate(year = as.character(year))
LLN_spec_unique <- LLN_spec_unique %>%
  mutate(year = as.character(year))


totalMass_year_per_plot <- aggLeafPlot_by_species %>%
    group_by(year, plot) %>%
    summarise(totalMass_year_plot = sum(mass_g_species, na.rm = TRUE), .groups = 'drop')

LLN_plotNew <- aggLeafPlot_by_species %>%
  inner_join(LLN_spec_unique, by = c("year", "plot", "week", "sci_name"),
             suffix = c("_litter", "_N_spec")) %>%
  left_join(aggwkpkPlot %>% select(year, plot, weekPeak, totalMass_year_plot),
            by = c("year", "plot")) %>%
  mutate(totalN_g = mass_g_species * (N_percent / 100))


####UGGG WE LOST OBSERVATIONS I'M GOING TO HAVE TO FIGURE THIS OUT
dim(LLN_plotNew)
dim(LLN_spec)
head(LLN_plotNew)

#TRYING TO TROUBLESHOOT BUT HAVEN'T DONE ANYTHING YET
aggLeafPlot_by_species_for_check <- aggLeafPlot_by_species %>%
  mutate(year = as.character(year)) # Make sure this matches your LLN_spec year type
lost_LLN_spec_rows <- LLN_spec_unique %>%
  anti_join(aggLeafPlot_by_species_for_check, by = c("year", "plot", "week", "sci_name"))
print(lost_LLN_spec_rows)



# Making a weighted average
total_mass_Nwksppltyr <- LLN_plot %>%
  group_by(year, plot, sci_name) %>%
  summarise(total_mass_species_in_N_weeks = sum(mass_g_species, na.rm = TRUE), .groups = 'drop')

LLN_plot_with_weighted_N <- LLN_plot %>%
  left_join(total_mass_Nwksppltyr, by = c("year", "plot", "sci_name")) %>%
  mutate(
    mass_prop_species_within_N_weeks = mass_g_species / total_mass_species_in_N_weeks,
    weighted_N_contribution_per_week = N_percent * mass_prop_species_within_N_weeks
  )

weighted_N_summary_per_species_plot_year <- LLN_plot_with_weighted_N %>%
  group_by(year, plot, sci_name) %>%
  summarise(
    weighted_avg_N_percent = sum(weighted_N_contribution_per_week, na.rm = TRUE),
    total_N_g_in_N_weeks = sum(totalN_g, na.rm = TRUE),
    total_mass_g_in_N_weeks = first(total_mass_species_in_N_weeks),
    .groups = 'drop'
  )

head(weighted_N_summary_per_species_plot_year)
dim(weighted_N_summary_per_species_plot_year)

LLN_peakNew <- LLN_plotNew %>%
  filter(week >= round(weekPeak) - 1 & week <= round(weekPeak) + 1)


# Now you have a new dataframe with only the samples you want
dim(LLN_peakNew)
head(LLN_peakNew)
```

```{r}
ggplot(data = LLN_peak, aes(x = weekPeak, y = N_percent, color=plot)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week Peak", y = "%N") +
  theme_minimal()

ggplot(data = LLN_plot, aes(x = week, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week", y = "%N") +
  theme_minimal()

ggplot(data = LLN_plot, aes(x = weekPeak, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week Peak", y = "%N") +
  theme_minimal()+
  xlim(42, 47)

ggplot(data = LLN_peak, aes(x = weekPeak, y = N_percent, color=sci_name)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week Peak", y = "%N") +
  theme_minimal()

ggplot(data = LLN_peak, aes(x = weekPeak, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week Peak", y = "%N") +
  theme_minimal()

ggplot(data = LLN_peak, aes(x = week, y = N_percent, color = as.factor(year))) + 
  facet_grid(~sci_name) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week", y = "%N", color = "year") +
  theme_minimal()

ggplot(data = LLN_peak, aes(x = weekPeak, y = N_percent, color = as.factor(year))) + 
  facet_grid(~sci_name) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week Peak", y = "%N", color = "year") +
  theme_minimal()
```

```{r}
ggplot(data = LLN_peak, aes(x = weekPeak, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week Peak", y = "%N") +
  theme_minimal()+
  xlim(42, 47)

ggplot(data = LLN_plot, aes(x = week, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week", y = "%N") +
  theme_minimal()

ggplot(data = LLN_peak, aes(x = weekPeak, y = totalN_g)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week Peak", y = "total N (g)") +
  theme_minimal()+
  xlim(42, 47)

ggplot(data = LLN_plot, aes(x = week, y = totalN_g)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Week", y = "total N (g)") +
  theme_minimal()
```


```{r}
# lmeNwkpk <- lme(weekPeak ~ N_percent, random=list(sci_name=~1, plot=~1), data=LLN_plot)
lmeNwkpk <- lme(N_percent ~ weekPeak, random=list(sci_name=~1, plot=~1), data=LLN_peak)
summary(lmeNwkpk)$tTable
anova(lmeNwkpk)

lmeNwkpkSpp <- lme(N_percent ~ weekPeak*sci_name, random=list(plot=~1), data=LLN_peak)
summary(lmeNwkpkSpp)$tTable
anova(lmeNwkpkSpp)

lmAll <- lme(N_percent ~ weekPeak, random=list(sci_name=~1, year=~1, week=~1, plot=~1), data=LLN_plot, na.action=na.omit)
summary(lmAll)$tTable

lmAllSPP <- lme(N_percent ~ weekPeak + sci_name, random=list(year=~1, week=~1, plot=~1), data=LLN_plot[,], na.action=na.omit)
anova(lmAllSPP)
summary(lmAllSPP)$tTable

lmAllACSA <- lme(N_percent ~ weekPeak, random=list(year=~1, week=~1, plot=~1), data=LLN_plot[LLN_plot$sci_name=="Acer saccharum",], na.action=na.omit)
summary(lmAllACSA)$tTable

lmAllQUAL <- lme(N_percent ~ weekPeak, random=list(year=~1, week=~1, plot=~1), data=LLN_plot[LLN_plot$sci_name=="Quercus alba",], na.action=na.omit)
summary(lmAllQUAL)$tTable


lmAll2 <- lme(N_percent ~ week, random=list(sci_name=~1, plot=~1), data=LLN_plot, na.action=na.omit)
summary(lmAll2)$tTable
```



```{r}
aggwkpkPlotWt <- leafTrapTotal %>%
  group_by(year, plot) %>%
  summarise(
    weekPeakWt = mean(weekPeakWt, na.rm = TRUE), 
    totalMass_year = first(totalMass_year),  
    .groups = 'drop'
  ) %>%
  mutate(year = as.character(year))
dim(leafTrapTotal)
dim(aggwkpkPlotWt)
head(aggwkpkPlotWt)

LLN_plotWt <- merge(aggwkpkPlotWt, LLN_spec, by = c("year", "plot"), all.x = TRUE)
##still something wrong with this
dim(LLN_plotWt)
dim(LLN_spec)
head(LLN_plotWt)

LLN_peakWt <- LLN_plotWt %>%
  filter(week >= round(weekPeakWt) - 0.1 & week <= round(weekPeakWt) + 0.1)
dim(LLN_peakWt)
head(LLN_peakWt)
```

```{r}
my_species_color_map <- c(
  "Quercus rubra" = "#8C3F48", # Muted Red/Pink
  "Quercus alba" = "#A6761D", # Muted Gold/Yellow
  "Acer saccharum" = "#006D6F"  # Light Purple
)

ggplot(data=LLN_peakWt, aes(x = weekPeakWt, y = N_percent)) +
  geom_smooth(method = "lm", se= FALSE, size = 2, color = title_color) + # Choose a single color
  geom_point(aes(color=sci_name), size = 3, alpha=0.7) +
  labs(
    x = "Week of Peak Leaf Drop",
    y = "Nitrogen Percentage\nin Leaf Litter (%)",
    color = "Species"
  ) + 
  scale_color_manual(values = my_species_color_map) + 
  theme_minimal(base_size = 20) +
 theme(
    axis.title = element_text(color = text_color),
    axis.text = element_text(color = text_color),
    legend.position = "none",
  )
```


