---
title: "07_Leaflitter_plottrap"
author: "Cierra Lizer"
date: "2025-07-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(lubridate)
library(dplyr)
library(readxl)
library(stringr)
library(readr)
library(nlme);  # Does the mixed effects model
library(emmeans) # will et us do a multi-comparisons test
```

```{r}
theme_base <-   theme(panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      panel.border = element_rect(fill="white", colour = "black", linewidth=0.7),
                      axis.title.x = element_text(margin = margin(t = 10, b=5), size=14),
                      axis.title.y = element_text(margin = margin(l = 5, r=5), size=14),
                      axis.text.x= element_text(margin = margin(t = 10), size=12),
                      axis.text.y=element_text(margin = margin(r = 10), size=12),
                      axis.ticks.length=unit(-0.3, "cm"),
                      # axis.ticks.margin=unit(0.5, "cm"),
                      axis.ticks = element_line(colour = "black", linewidth = 0.4))


# Setting a consistent color scheme across all graphs
plotOrder <- c("B-127", "U-134", "N-115", "HH-115")
ewPlotColors <- c("#1B9E77","#D95F02", "#7570B3", "#E7298A")
names(ewPlotColors) = plotOrder
ewPlotColors
```

```{r}
path.google <- "~/Google Drive/My Drive"
path.litter <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data")
path.CN <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_runs")
path.figs <- file.path(path.litter, "figures") # where we shoudl save some figures
path.save <- file.path(path.litter, "LeafLitterData_Clean_forArchiving") # Where we shoudl save the data
cierra.path <- "G:/.shortcut-targets-by-id/0B_Fbr697pd36TkVHdDNJQ1dJU1E/East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_Runs"
```

___________________________________________________________________________________________________________________________________________

*Cleaning Leaf litter*

```{r}
#combine into one dataset
all_cleaned_files <- list.files(path.save, pattern = "\\.csv$", full.names = TRUE)
if (length(all_cleaned_files) == 0) { stop(...) }
datLitter <- map_df(all_cleaned_files, read_csv)
datLitter$plot <- factor(datLitter$plot, levels = plotOrder)
summary(datLitter)

datLitter$year <- lubridate::year(datLitter$date_collection)
datLitter$yday <- lubridate::yday(datLitter$date_collection)
datLitter$week <- lubridate::week(datLitter$date_collection)
summary(datLitter)

datLitter <- datLitter[datLitter$year<2024,]
summary(datLitter)
```

Cleaning of missing and empty's
```{r}
# There is no HH-115 NE, NW, SE, SW --> it has a weird layout
#datLitter[datLitter$plot=="HH-115" & datLitter$trap_ID %in% c("NE", "NW", "SE", "SW"), "trap_ID"] <- NA

datLitter <- datLitter[!is.na(datLitter$trap_ID),]

# Finding bags that that are empty or missing
unique(datLitter$tissue)

datMissing <- datLitter[datLitter$tissue == "MISSING DATA",]
dim(datMissing)
summary(datMissing)

datNone <- datLitter[datLitter$tissue == "EMPTY BAG",]
dim(datNone)
summary(datNone)
```

Summing to the trap level --> good to use for comparisons of plots and forest wide things
```{r}
aggLeafTrap <- aggregate(mass_g ~ year + week + date_collection + plot + trap_ID, data=datLitter[datLitter$tissue=="leaf",], FUN=sum)
summary(aggLeafTrap)

aggLeafTrap <- rbind(aggLeafTrap, datNone[,c("year", "week", "date_collection", "plot", "trap_ID", "mass_g")], datMissing[,c("year", "week", "date_collection", "plot", "trap_ID", "mass_g")])
summary(aggLeafTrap)
```

```{r}
allCollect = NULL
# plotsCore <- c("N", "S", "E", "W")
# plotExtra <- c("NW", "SW", "NE", "SE")
for(PLT in unique(aggLeafTrap$plot)){
  datPLT <- aggLeafTrap[aggLeafTrap$plot==PLT,]
  datPLT <- droplevels(datPLT)
  
  datePLT <- unique(datPLT$date_collection)
  datePLT <- sort(datePLT)
  daysPLT <- diff(datePLT)
  # Making a data frame with all collections dates and traps for the plot
  # That way we can add in plots as missing if they're not there
  dfCollection <- data.frame(plot=PLT, trap_ID = rep(unique(datPLT$trap_ID), each=length(datePLT)),
                             date_collection=datePLT, days_collection=c(NA, daysPLT))
                            
  
  if(is.null(allCollect)){
    allCollect <- dfCollection
  } else {
    allCollect <- rbind(allCollect, dfCollection)
  }
}
summary(allCollect)

# Merging the number of days in each collection period with the actual data

dim(aggLeafTrap)
aggLeafTrap <- merge(aggLeafTrap, allCollect, all=T)
aggLeafTrap$year <- lubridate::year(aggLeafTrap$date_collection)
aggLeafTrap$week <- lubridate::week(aggLeafTrap$date_collection)
summary(aggLeafTrap)

# This added in a lot of NAs because we added half our traps a couple years in
for(PLT in unique(aggLeafTrap$plot)){
  for(TRP in unique(aggLeafTrap$trap_ID[aggLeafTrap$plot==PLT])){
    minTrap <- min(aggLeafTrap$date_collection[aggLeafTrap$plot==PLT & aggLeafTrap$trap_ID==TRP & !is.na(aggLeafTrap$mass_g)] ,na.rm=T)
    
    # test <- aggLeafTrap[aggLeafTrap$plot==PLT & aggLeafTrap==TRP & aggLeafTrap$date_collection>minTrap,]
    
    aggLeafTrap <- aggLeafTrap[aggLeafTrap$plot!=PLT | (aggLeafTrap$plot==PLT & aggLeafTrap$trap_ID!=TRP) | 
                                 (aggLeafTrap$plot==PLT & aggLeafTrap$trap_ID==TRP & aggLeafTrap$date_collection>minTrap),]
  }
}


dim(aggLeafTrap)
summary(aggLeafTrap)

aggLeafTrap$mass_g_day <- aggLeafTrap$mass_g/aggLeafTrap$days_collection
summary(aggLeafTrap)
```

```{r}
# Lets calculate the proportion of leaf fall at any given point in time
leafTrapTotal <- aggregate(mass_g ~ year + plot + trap_ID, data=aggLeafTrap, FUN=sum)
names(leafTrapTotal)[names(leafTrapTotal)=="mass_g"] <- "totalMass_year"
summary(leafTrapTotal)
summary(leafTrapTotal[leafTrapTotal$year<2024,])

# Merging our totals into the weekly sums so we can get proportion
aggLeafTrap <- merge(aggLeafTrap, leafTrapTotal, all=T)
aggLeafTrap <- aggLeafTrap[!is.na(aggLeafTrap$trap_ID),]
aggLeafTrap$mass_prop <- aggLeafTrap$mass_g/aggLeafTrap$totalMass_year
summary(aggLeafTrap)
aggLeafTrap[is.na(aggLeafTrap$totalMass_year),]
```

Plot level --> Will need for nitrogen analysis
```{r}
# Getting some plot-level summary stats
aggLeafPlot <- aggregate(cbind(mass_g, mass_g_day, mass_prop, totalMass_year)~ year + plot + week + date_collection, data=aggLeafTrap, FUN=mean)
summary(aggLeafPlot)
```

```{r}
weekPeak <- data.frame(year=rep(unique(aggLeafTrap$year), each=length(unique(aggLeafPlot$plot))),
                       plot=rep(unique(aggLeafPlot$plot)),
                       week=NA,
                       date=NA,
                       propPeak = NA,
                       prop9Wk = NA,
                       prop5Wk = NA,
                       prop3Wk = NA)

for(YR in unique(weekPeak$year)){
  datYr <- aggLeafPlot[aggLeafPlot$year==YR,]
  for(PLT in unique(datYr$plot)){
    indNow <- which(weekPeak$year==YR & weekPeak$plot==PLT)
    datPlot <- datYr[datYr$plot==PLT,]
    
    indPeak <- which(datPlot$mass_g_day==max(datPlot$mass_g_day))
    wkPeak <- datPlot$week[indPeak]
    weekPeak$week[indNow] <- wkPeak
    weekPeak$date[indNow] <- as.character(datPlot$date_collection[indPeak])
    weekPeak$propPeak[indNow] <- datPlot$mass_prop[indPeak]
    weekPeak$prop9Wk[indNow] <- sum(datPlot$mass_prop[datPlot$week %in% (wkPeak-4):(wkPeak + 4)])
    weekPeak$prop5Wk[indNow] <- sum(datPlot$mass_prop[datPlot$week %in% (wkPeak-2):(wkPeak + 2)])
    weekPeak$prop3Wk[indNow] <- sum(datPlot$mass_prop[datPlot$week %in% (wkPeak-1):(wkPeak + 1)])
  }
}
weekPeak$date <- as.Date(weekPeak$date)
summary(weekPeak)
weekPeak[weekPeak$prop5Wk<0.5,]

#write.csv(weekPeak, file.path(path.google, "URF REU 2025 - Lizer - Leaf Litter ", "PeakLeafDates_byPlot.csv"), row.names=F)
```

```{r}
summary(leafTrapTotal)
for(YR in unique(aggLeafTrap$year)){
  print(YR)
  datYr <- aggLeafTrap[aggLeafTrap$year==YR,]
  for(PLT in unique(datYr$plot)){
    print(PLT)
    datPlot <- datYr[datYr$plot==PLT,]
    for(TRP in unique(datPlot$trap_ID)){
      indNow <- which(leafTrapTotal$year==YR & leafTrapTotal$plot==PLT & leafTrapTotal$trap_ID==TRP)
      datTrp <- datPlot[datPlot$trap_ID==TRP,]
  
      print(paste(TRP, "-", indNow))
      indPeak <- which(datTrp$mass_g_day==max(datTrp$mass_g_day, na.rm=T))
      wkPeak <- datTrp$week[indPeak]
      # tmpPleafTrapTotaleak <- data.frame(year = YR, plot=PLT, trap_ID=TRP, week=NA, date=NA, propPeak=NA)
      leafTrapTotal[indNow,"weekPeak"] <- wkPeak
      leafTrapTotal[indNow,"date"] <- as.character(datTrp$date_collection[indPeak])
      leafTrapTotal[indNow,"propPeak"] <- datTrp$mass_prop[indPeak]
      # tmpPeak[1,"prop9Wk"] <- sum(datTrp$mass_prop[datTrp$week %in% (wkPeak-4):(wkPeak + 4)])
      # tmpPeak[1,"prop5Wk"] <- sum(datTrp$mass_prop[datTrp$week %in% (wkPeak-2):(wkPeak + 2)])
      # tmpPeak[1,"prop3Wk"] <- sum(datTrp$mass_prop[datTrp$week %in% (wkPeak-1):(wkPeak + 1)])
    }
  }
}
# weekPeakTrap$date <- as.Date(weekPeakTrap$date)
summary(leafTrapTotal)
summary(leafTrapTotal[is.na(leafTrapTotal$weekPeak),])
# weekPeak[weekPeak$prop5Wk<0.5,]
```

___________________________________________________________________________________________________________________________________________

*LL & Met*
Want it aggregated at the trap level so we can have replicates within plots

```{r}
metSummer <- read.csv("G:/.shortcut-targets-by-id/1q2wvODXrDo0tgOTLpFqF7TqcWoKoHZjW/URF REU 2025 - Lizer - Leaf Litter/data/daymet/daymet_June-July-August_summaries_2017-2023.csv")
```

```{r}
leafTrapMet <- merge(leafTrapTotal, metSummer, all.x=T)
summary(leafTrapMet)
```

```{r}
lmeYrs <- lme(weekPeak ~ as.factor(year), random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeYrs)
anova(lmeYrs)

lmeLLRainless <- lme(weekPeak ~ n.Rainless, random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLRainless)
anova(lmeLLRainless)

sumlmeLLRainlessConsec <- lme(weekPeak ~ RainlessConsec.max , random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLRainlessConsec)
anova(lmeLLRainlessConsec)

lmeLLPrcp <- lme(weekPeak ~ prcp..mm.day., random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLPrcp)
anova(lmeLLPrcp)

lmeLLTmax <- lme(weekPeak ~ tmax..deg.c., random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLTmax)$tTable
anova(lmeLLTmax)

lmeLLVP <- lme(weekPeak ~ vp..Pa., random=list(plot=~1, trap_ID=~1), data=leafTrapMet)
summary(lmeLLVP)
anova(lmeLLVP)
```

```{r}
ggplot(data=leafTrapMet) +
  geom_boxplot(aes(x=as.factor(year), y=weekPeak)) +
  geom_point(aes(x=as.factor(year), y=weekPeak, color=plot), position=position_jitter(0.1))


ggplot(data=leafTrapMet) +
  # geom_boxplot(aes(x=as.factor(year), y=weekPeak)) +
  geom_point(aes(x=n.Rainless, y=weekPeak, color=plot))

ggplot(data=leafTrapMet) +
  # geom_boxplot(aes(x=as.factor(year), y=weekPeak)) +
  geom_point(aes(x=prcp..mm.day., y=weekPeak, color=plot))

ggplot(data=leafTrapMet)+
  geom_point(aes(x=RainlessConsec.max, y=weekPeak, color=plot))

ggplot(data=leafTrapMet)+
  geom_point(aes(x=tmax..deg.c., y=weekPeak, color=plot))

ggplot(data=leafTrapMet)+
  geom_point(aes(x=vp..Pa., y=weekPeak, color=plot))
```

___________________________________________________________________________________________________________________________________________

*Cleaning N data*
Loading in data from 2021-22
```{r}
LLN2122path<-list.files(cierra.path, pattern = "Lizer_EWLL_2021-2022_run.xlsx", full.names = TRUE)
LLN2122 <- read_excel(LLN2122path)
head(LLN2122)
```

Name (aka sample id) will be composed of plot (first 4-5 characters), underscore, date (next 8 characters), underscore, species id (last 4 characters). Date is in the form year (4 characters), month (2 characters), day (2 characters)

Loading in 2018 data
```{r}
LLN18path<-list.files(cierra.path, pattern = "Fall_2018_run1.csv", full.names = TRUE)
LLN18 <- read_csv(LLN18path)
head(LLN18)
```
Name is composed of plot (1st character), dash, year (2 characters), month (2 character), day (2 characters), species id (2 characters)

Rename columns to be easier to read and pull information out from the name
```{r}
LLN2122_clean <- LLN2122 %>%
  rename(
    CN_ratio = `C/N  ratio`,
    N_percent = `N  [%]`,
    C_percent = `C  [%]`
  ) %>%
  
  mutate(
    spec_id = str_sub(Name, -4), # Get the last 4 characters
    date_raw = str_sub(Name, -13, -5), #finds the 8 characters before the last underscore (5th to last character)
    #note: will give year, month, day in a string together
    plot = str_extract(Name, "^[^_]+") # Extracts characters from start (^) until the first underscore (_)
  ) %>%
#convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, everything(), -date_raw # Keep original, then new columns, then rest
  )

print(LLN2122_clean)
```

A BUNCH of cleaning 2018 and making the formats similar to what I had done
```{r}
LLN18_clean <- LLN18 %>%
  rename(
    CN_ratio = `C:N`,
    N_percent = `%N`,
    C_percent = `%C`
  ) %>%
  mutate(
    spec_id_raw = str_sub(Name, 11,12),
    year_2dig = str_sub(Name, 3, 4),
    month_2dig = str_sub(Name, 6,7),
    day_2dig = str_sub(Name, 9,10),
    plot_raw = str_sub(Name, 1,1)
  ) %>%
  mutate(
    spec_id = case_when(
      spec_id_raw == "QA" ~ "QUAL",
      spec_id_raw == "TA" ~ "TIAM",
      spec_id_raw == "AS" ~ "ACSA",
      TRUE ~ spec_id_raw
    ), 
    plot = case_when(
      plot_raw == "H" ~ "HH115",
      plot_raw == "N" ~ "N115",
      plot_raw == "U" ~ "U134",
      plot_raw == "B" ~ "B127",
      TRUE ~ plot_raw
    )
  ) %>% 
  mutate(
    date_raw = paste0("20", year_2dig, month_2dig, day_2dig)
  ) %>%
  # convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, CN_ratio, N_percent, C_percent, everything(),
    -year_2dig, -month_2dig, -day_2dig, -date_raw, -plot_raw, -spec_id_raw
  )

print(LLN18_clean)
unique(LLN18_clean$spec_id)
unique(LLN18_clean$plot)
```

AGGGGHHH I accidentally put QUAB and QUAL both meaning Quercus alba
```{r}
LLN2122_clean <- LLN2122_clean %>%
  mutate(
    spec_id = case_when(
    spec_id == "QUAB" ~ "QUAL", # If spec_id is "QUAB", change it to "QUAL"
    TRUE ~ spec_id
  )
  )


LLN2122_clean <- LLN2122_clean %>%
  mutate(
    plot = case_when(
    plot == "H115" ~ "HH115",
    TRUE ~ plot
    )
  )

unique(LLN2122_clean$spec_id)
unique(LLN2122_clean$plot)
```

Combining my 2 dataframes
```{r}
LLNdattot <- bind_rows(LLN2122_clean, LLN18_clean)

dim(LLNdattot)
unique(LLNdattot$year)
```

Removing Tilia Americana from the dataset
```{r}
LLN_spec <- LLNdattot %>%
  filter(spec_id != "TIAM")

unique(LLN_spec$spec_id)
dim(LLN_spec)
```
```{r}
#write.csv(LLN_spec, file.path(path.google, "URF REU 2025 - Lizer - Leaf Litter ", "CNcombined.csv"), row.names=F)
```

```{r}
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "QUAL" ~ "Quercus alba", 
    TRUE ~ spec_id
  )
  )
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "QURU" ~ "Quercus rubra", 
    TRUE ~ sci_name
  )
  )
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "ACSA" ~ "Acer Saccharum", 
    TRUE ~ sci_name
  )
  )

print(LLN_spec)
```
___________________________________________________________________________________________________________________________________________
*Looking at just N*
```{r}
ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = plot)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week", y = "C/N Ratio", color = "plot") +
  theme_minimal()

ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = plot)) + 
  facet_grid(~sci_name) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week", y = "C/N Ratio", color = "plot") +
  theme_minimal()
```

```{r}
lmeNyear <- lme(N_percent ~ as.factor(year), random=list(spec_id=~1, plot=~1, week=~1), data=LLN_spec)
summary(lmeNyear)
anova(lmeNyear)
```

___________________________________________________________________________________________________________________________________________
*N and Met data*
```{r}
NMet <- merge(LLN_spec, metSummer,  by = c("year"), all.x=T)
summary(NMet)
```

```{r}
ggplot(data = NMet, aes(x = n.Rainless, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Number of Rainless Days", y = "%N", color = "Species") +
  theme_minimal()

ggplot(data = NMet, aes(x = prcp..mm.day., y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Number of Rainless Days", y = "%N", color = "Species") +
  theme_minimal()

ggplot(data = NMet, aes(x = RainlessConsec.max, y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Number of Consecutive Rainless Days", y = "%N") +
  theme_minimal()

ggplot(data = NMet, aes(x = tmax..deg.c., y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Maximum Temperature", y = "%N") +
  theme_minimal()

ggplot(data = NMet, aes(x = vp..Pa., y = N_percent)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Vapor Pressure", y = "%N") +
  theme_minimal()
```
```{r}
lmeNRainless <- lme(n.Rainless ~ N_percent, random=list(plot=~1), data=NMet)
summary(lmeNRainless)
anova(lmeNRainless)

lmeNRainlessConsec <- lme(RainlessConsec.max ~ N_percent, random=list(plot=~1), data=NMet)
summary(lmeNRainlessConsec)
anova(lmeNRainlessConsec)

lmeNPrcp <- lme(prcp..mm.day. ~ N_percent, random=list(plot=~1), data=NMet)
summary(lmeNPrcp)
anova(lmeNPrcp)

lmeNTmax <- lme(tmax..deg.c. ~ N_percent, random=list(plot=~1), data=NMet)
summary(lmeNTmax)
anova(lmeNTmax)

lmeNVP <- lme(vp..Pa. ~ N_percent, random=list(plot=~1), data=NMet)
summary(lmeNVP)
anova(lmeNVP)
```


___________________________________________________________________________________________________________________________________________

*N and LL data*
Making sure N and LL data align
```{r}
unique(LLN_spec$plot)
unique(leafTrapTotal$plot)

LLN_spec <- LLN_spec %>%
  mutate(plot = recode(plot,
  "B127" = "B-127",
  "N115" = "N-115",
  "HH115" = "HH-115",
  "U134" = "U-134"
  ))

unique(LLN_spec$plot)
```


```{r}
aggwkpkPlot <- aggregate(weekPeak ~ year + plot, data = leafTrapTotal, FUN = mean)
aggwkpkPlot$year <- as.character(aggwkpkPlot$year)

summary(aggwkpkPlot)

LLN_plot <- merge(aggwkpkPlot, LLN_spec, by = c("year", "plot"), all.x = TRUE)

LLN_peak <- LLN_plot %>%
  filter(week == round(weekPeak))

# Now you have a new dataframe with only the samples you want
dim(LLN_peak)
head(LLN_peak)

```

```{r}
ggplot(data = LLN_peak, aes(x = weekPeak, y = N_percent, color=plot)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week Peak", y = "%N") +
  theme_minimal()
```

```{r}
# lmeNwkpk <- lme(weekPeak ~ N_percent, random=list(sci_name=~1, plot=~1), data=LLN_plot)
lmeNwkpk <- lme(N_percent ~ weekPeak, random=list(sci_name=~1, plot=~1), data=LLN_peak)
summary(lmeNwkpk)$tTable
anova(lmeNwkpk)

lmeNwkpkSpp <- lme(N_percent ~ weekPeak*sci_name, random=list(plot=~1), data=LLN_peak)
summary(lmeNwkpkSpp)$tTable
anova(lmeNwkpkSpp)

```

