---
title: "08_Leaflitter_species"
author: "Cierra Lizer"
date: "2025-07-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(lubridate)
library(dplyr)
library(readxl)
library(stringr)
library(readr)
library(nlme);  # Does the mixed effects model
library(emmeans) # will et us do a multi-comparisons test
```

```{r}
theme_base <-   theme(panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      panel.border = element_rect(fill="white", colour = "black", linewidth=0.7),
                      axis.title.x = element_text(margin = margin(t = 10, b=5), size=14),
                      axis.title.y = element_text(margin = margin(l = 5, r=5), size=14),
                      axis.text.x= element_text(margin = margin(t = 10), size=12),
                      axis.text.y=element_text(margin = margin(r = 10), size=12),
                      axis.ticks.length=unit(-0.3, "cm"),
                      # axis.ticks.margin=unit(0.5, "cm"),
                      axis.ticks = element_line(colour = "black", linewidth = 0.4))


# Setting a consistent color scheme across all graphs
plotOrder <- c("B-127", "U-134", "N-115", "HH-115")
ewPlotColors <- c("#1B9E77","#D95F02", "#7570B3", "#E7298A")
names(ewPlotColors) = plotOrder
ewPlotColors
```

```{r}
path.google <- "~/Google Drive/My Drive"
path.litter <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data")
path.CN <- file.path(path.google, "East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_runs")
path.figs <- file.path(path.litter, "figures") # where we shoudl save some figures
path.save <- file.path(path.litter, "LeafLitterData_Clean_forArchiving") # Where we shoudl save the data
cierra.path <- "G:/.shortcut-targets-by-id/0B_Fbr697pd36TkVHdDNJQ1dJU1E/East Woods/Rollinson_Monitoring/Data/Leaf_litter_data/CN_Runs"
```

___________________________________________________________________________________________________________________________________________

*Cleaning Leaf litter*

```{r}
#combine into one dataset
all_cleaned_files <- list.files(path.save, pattern = "\\.csv$", full.names = TRUE)
if (length(all_cleaned_files) == 0) { stop("No cleaned leaf litter files found in path.save.") }
datLitter <- map_df(all_cleaned_files, read_csv)
datLitter$plot <- factor(datLitter$plot, levels = plotOrder)
summary(datLitter)

datLitter$year <- lubridate::year(datLitter$date_collection)
datLitter$yday <- lubridate::yday(datLitter$date_collection)
datLitter$week <- lubridate::week(datLitter$date_collection)
summary(datLitter)

datLitter <- datLitter[datLitter$year<2024,]
summary(datLitter)
```

Cleaning of missing and empty's
```{r}
# There is no HH-115 NE, NW, SE, SW --> it has a weird layout
#datLitter[datLitter$plot=="HH-115" & datLitter$trap_ID %in% c("NE", "NW", "SE", "SW"), "trap_ID"] <- NA

datLitter <- datLitter[!is.na(datLitter$trap_ID),]

# Finding bags that that are empty or missing
unique(datLitter$tissue)

datMissing <- datLitter[datLitter$tissue == "MISSING DATA",]
dim(datMissing)
summary(datMissing)

datNone <- datLitter[datLitter$tissue == "EMPTY BAG",]
dim(datNone)
summary(datNone)
```

Combining species and trap
```{r}
spectrp <- bind_rows(
  datLitter %>%
    filter(tissue == "leaf") %>%
    select(year, week, date_collection, plot, trap_ID, mass_g, genus, species),
  datNone %>%
    select(year, week, date_collection, plot, trap_ID, mass_g, genus, species),
  datMissing %>%
    select(year, week, date_collection, plot, trap_ID, mass_g, genus, species)
) %>%
  mutate(sci_name = paste(genus, species)) %>%
  group_by(year, week, date_collection, plot, trap_ID) %>%
  mutate(mass_g_trap_total = sum(mass_g, na.rm = TRUE)) %>%
  ungroup()
```

**PROBLEM IN THIS CHUNK**
Error in data.frame(plot = PLTsp, trap_ID = rep(unique(datPLTsp$trap_ID),  : 
  arguments imply differing number of rows: 1, 0
```{r}
allCollectSp = NULL
for(PLTsp in unique(spectrp$plot)){ 
  datPLTsp <- spectrp[spectrp$plot==PLTsp,]
  datPLTsp <- droplevels(datPLTsp)
  
  datePLTsp <- unique(datPLTsp$date_collection)
  datePLTsp <- sort(datePLTsp)
  daysPLTsp <- diff(datePLTsp)
  
  dfCollectionsp <- data.frame(plot=PLTsp, trap_ID = rep(unique(datPLTsp$trap_ID), each=length(datePLTsp)),
                             date_collection=datePLTsp, days_collection=c(NA, daysPLTsp))
  
  if(is.null(allCollectSp)){
    allCollectSp <- dfCollectionsp
  } else {
    allCollectSp <- rbind(allCollectSp, dfCollectionsp)
  }
}
summary(allCollectSp)

#creating equivalent of aggleaftrap
datLitter_sptp <- merge(spectrp, allCollectSp, all=T)
datLitter_sptp$year <- lubridate::year(datLitter_sptp$date_collection)
datLitter_sptp$week <- lubridate::week(datLitter_sptp$date_collection)

# 6. Filter out pre-deployment NA rows
for(PLTsp in unique(datLitter_sptp$plot)){
  for(TRPsp in unique(datLitter_sptp$trap_ID[datLitter_sptp$plot==PLTsp])){
    minTrapsp <- min(datLitter_sptp$date_collection[datLitter_sptp$plot==PLTsp & datLitter_sptp$trap_ID==TRPsp & !is.na(datLitter_sptp$mass_g)] ,na.rm=T)
    datLitter_sptp <- datLitter_sptp[datLitter_sptp$plot!=PLTsp | (datLitter_sptp$plot==PLTsp & datLitter_sptp$trap_ID!=TRPsp) | 
                                               (datLitter_sptp$plot==PLTsp & datLitter_sptp$trap_ID==TRPsp & datLitter_sptp$date_collection >= minTrapsp),]
  }

```


Calculate mass rates and proportions at both species in trap and total trap levels
```{r}
datLitter_sptp$mass_g_day_sp <- datLitter_sptp$mass_g / datLitter_sptp$days_collection # Species-specific mass per day
datLitter_sptp$mass_g_day_trp <- datLitter_sptp$mass_g_trp / datLitter_sptp$days_collection # Total trap mass per day
```

Calculate proportions of total annual trap mass
```{r}
datLitter_sptp <- datLitter_sptp %>%
  group_by(year, plot, trap_ID) %>%
  # Calculate total annual mass per trap
  mutate(totalMass_trap_year = sum(mass_g_trp, na.rm = TRUE)) %>% 
  ungroup() %>%
  # Proportion of indv species mass relative to total annual trap mass
  mutate(mass_prop_spec_in_trap = mass_g / totalMass_trap_year) %>%
  # Proportion of total trap mass at each collection relative to total annual trap mass
  mutate(mass_prop_trap_total = mass_g_trp / totalMass_trap_year)

summary(datLitter_sptp)
```

___________________________________________________________________________________________________________________________________________

*Looking at just LL dat*
```{r}
ggplot(data=aggLeafSppTot[aggLeafSppTot$sci_name %in% c("Acer saccharum", "Quercus alba", "Quercus rubra") & aggLeafSppTot$year < 2023,]) +
  facet_wrap(~sci_name, ncol=2) +
  geom_point(aes(x=as.factor(year), y=weekPeak, color=plot), position=position_jitter(0.1)) +
  geom_boxplot(aes(x=as.factor(year), y=weekPeak)) 
```

```{r}
lmeYrsSpp <- lme(weekPeak ~ as.factor(year)*sci_name, random=list(plot=~1, trap_ID=~1), data=aggLeafSppTot[aggLeafSppTot$sci_name %in% c("Acer saccharum", "Quercus alba", "Quercus rubra"),])
summary(lmeYrsSpp)
anova(lmeYrsSpp)


lmeYrsQURU <- lme(weekPeak ~ as.factor(year), random=list(plot=~1, trap_ID=~1), data=aggLeafSppTot[aggLeafSppTot$sci_name %in% c("Quercus rubra"),])
summary(lmeYrsQURU)
anova(lmeYrsQURU)

quruComp <- emmeans(lmeYrsQURU, ~year)
pairs(quruComp, adjust="tukey")


lmeYrsQUAL <- lme(weekPeak ~ as.factor(year), random=list(plot=~1, trap_ID=~1), data=aggLeafSppTot[aggLeafSppTot$sci_name %in% c("Quercus alba"),])
summary(lmeYrsQUAL)
anova(lmeYrsQUAL)

qualComp <- emmeans(lmeYrsQUAL, ~year)
pairs(qualComp, adjust="tukey")


lmeYrsACSA <- lme(weekPeak ~ as.factor(year), random=list(plot=~1, trap_ID=~1), data=aggLeafSppTot[aggLeafSppTot$sci_name %in% c("Acer saccharum"),])
summary(lmeYrsACSA)
anova(lmeYrsACSA)

acsaComp <- emmeans(lmeYrsACSA, ~year)
pairs(acsaComp, adjust="tukey")
```

___________________________________________________________________________________________________________________________________________

*LL & Met*

___________________________________________________________________________________________________________________________________________

*Cleaning N data*
Loading in data from 2021-22
```{r}
LLN2122path<-list.files(cierra.path, pattern = "Lizer_EWLL_2021-2022_run.xlsx", full.names = TRUE)
LLN2122 <- read_excel(LLN2122path)
head(LLN2122)
```

Name (aka sample id) will be composed of plot (first 4-5 characters), underscore, date (next 8 characters), underscore, species id (last 4 characters). Date is in the form year (4 characters), month (2 characters), day (2 characters)

Loading in 2018 data
```{r}
LLN18path<-list.files(cierra.path, pattern = "Fall_2018_run1.csv", full.names = TRUE)
LLN18 <- read_csv(LLN18path)
head(LLN18)
```
Name is composed of plot (1st character), dash, year (2 characters), month (2 character), day (2 characters), species id (2 characters)

Rename columns to be easier to read and pull information out from the name
```{r}
LLN2122_clean <- LLN2122 %>%
  rename(
    CN_ratio = `C/N  ratio`,
    N_percent = `N  [%]`,
    C_percent = `C  [%]`
  ) %>%
  
  mutate(
    spec_id = str_sub(Name, -4), # Get the last 4 characters
    date_raw = str_sub(Name, -13, -5), #finds the 8 characters before the last underscore (5th to last character)
    #note: will give year, month, day in a string together
    plot = str_extract(Name, "^[^_]+") # Extracts characters from start (^) until the first underscore (_)
  ) %>%
#convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, everything(), -date_raw # Keep original, then new columns, then rest
  )

print(LLN2122_clean)
```

A BUNCH of cleaning 2018 and making the formats similar to what I had done
```{r}
LLN18_clean <- LLN18 %>%
  rename(
    CN_ratio = `C:N`,
    N_percent = `%N`,
    C_percent = `%C`
  ) %>%
  mutate(
    spec_id_raw = str_sub(Name, 11,12),
    year_2dig = str_sub(Name, 3, 4),
    month_2dig = str_sub(Name, 6,7),
    day_2dig = str_sub(Name, 9,10),
    plot_raw = str_sub(Name, 1,1)
  ) %>%
  mutate(
    spec_id = case_when(
      spec_id_raw == "QA" ~ "QUAL",
      spec_id_raw == "TA" ~ "TIAM",
      spec_id_raw == "AS" ~ "ACSA",
      TRUE ~ spec_id_raw
    ), 
    plot = case_when(
      plot_raw == "H" ~ "HH115",
      plot_raw == "N" ~ "N115",
      plot_raw == "U" ~ "U134",
      plot_raw == "B" ~ "B127",
      TRUE ~ plot_raw
    )
  ) %>% 
  mutate(
    date_raw = paste0("20", year_2dig, month_2dig, day_2dig)
  ) %>%
  # convert date_raw into a readable year format
  mutate(
    date = ymd(date_raw),
    year = year(date),
    week = isoweek(date)
  ) %>%
  mutate(year = factor(year)) %>%
  select(
    Name, plot, date, year, week, spec_id, CN_ratio, N_percent, C_percent, everything(),
    -year_2dig, -month_2dig, -day_2dig, -date_raw, -plot_raw, -spec_id_raw
  )

print(LLN18_clean)
unique(LLN18_clean$spec_id)
unique(LLN18_clean$plot)
```

AGGGGHHH I accidentally put QUAB and QUAL both meaning Quercus alba
```{r}
LLN2122_clean <- LLN2122_clean %>%
  mutate(
    spec_id = case_when(
    spec_id == "QUAB" ~ "QUAL", # If spec_id is "QUAB", change it to "QUAL"
    TRUE ~ spec_id
  )
  )


LLN2122_clean <- LLN2122_clean %>%
  mutate(
    plot = case_when(
    plot == "H115" ~ "HH115",
    TRUE ~ plot
    )
  )

unique(LLN2122_clean$spec_id)
unique(LLN2122_clean$plot)
```

Combining my 2 dataframes
```{r}
LLNdattot <- bind_rows(LLN2122_clean, LLN18_clean)

dim(LLNdattot)
unique(LLNdattot$year)
```

Removing Tilia Americana from the dataset
```{r}
LLN_spec <- LLNdattot %>%
  filter(spec_id != "TIAM")

unique(LLN_spec$spec_id)
dim(LLN_spec)
```

```{r}
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "QUAL" ~ "Quercus alba", 
    TRUE ~ sci_name
  )
  )
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "QURU" ~ "Quercus rubra", 
    TRUE ~ sci_name
  )
  )
LLN_spec <- LLN_spec %>%
  mutate(
    sci_name = case_when(
    spec_id == "ACSA" ~ "Acer Saccharum", 
    TRUE ~ sci_name
  )
  )
```

___________________________________________________________________________________________________________________________________________

*Looking at just N*
```{r}
ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = spec_id)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week", y = "C/N Ratio", color = "Species") +
  theme_minimal()

ggplot(data = LLN_spec, aes(x = week, y = N_percent, color = as.factor(year))) +
   facet_grid(~spec_id) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Week", y = "N_percent", color = "Year") +
  theme_minimal()
```

```{r}
lmeNyear <- lme(N_percent ~ as.factor(year), random=list(spec_id=~1, plot=~1, week=~1), data=LLN_spec)
summary(lmeNyear)
anova(lmeNyear)
```
__________________________________________________________________________________________________________________________________________

*N and Met data*

```{r}
NMet <- merge(LLN_spec, metSummer, all.x=T)
summary(NMet)
```

```{r}
ggplot(data = NMet, aes(x = n.Rainless, y = N_percent, color = sci_name)) + 
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Number of Rainless Days", y = "%N", color = "Species") +
  theme_minimal()
```

___________________________________________________________________________________________________________________________________________

*N and LL data*
